---
title: "Sample Sheet Report"
author: 'Nick Dillen'
date: "`r Sys.Date()`"
output:
  pdf_document: 
    toc: true
    fig_caption: true
  html_document:
    code_folding: hide
    highlight: haddock
    number_sections: false
    fig_caption: true
    theme: sandstone
    toc: true
    toc_float: true
urlcolor: blue
params: 
  SAMPLE_SHEET_URL: "SAMPLE_SHEET_URL"
  BC_OUT_DIR: "BC_OUT_DIR"
---

```{r user-input, echo=F}
library(knitr)
library(gendiver)

SS_URL = params$SAMPLE_SHEET_URL

print(params)

```


```{r gs4-auth, message=FALSE, echo=F}
googlesheets4::gs4_auth(T)

ss.df = data.frame(googlesheets4::read_sheet(SS_URL))

# Remove NA rows
ss.df[ss.df == "#N/A"] = NA
```



\break

# `r unique(ss.df$run[!is.na(ss.df$run)])`

Used sample sheet link:

- `r knitr::asis_output(SS_URL)`

## INFO: number of rows with USI

- Number of rows in sample_sheet **total**: `r nrow(ss.df)`
- Number of rows in sample_sheet **with** unique_sample_code field: `r sum(!is.na(ss.df$unique_sample_code))`
- Number of rows in sample_sheet **without** unique_sample_code field: `r sum(is.na(ss.df$unique_sample_code))`

```{r rm-NA}
ss.df = ss.df[!is.na(ss.df$unique_sample_code),]
```

## CHECK: are the USI codes unique?

USI = `unique_sample_code`

- Number of PCR-amplicons/rows with **duplicated** USI: `r sum(duplicated(ss.df$unique_sample_code))`
- Number of PCR-amplicons/rows with **unique** USI: `r sum(!duplicated(ss.df$unique_sample_code))`


```{r USI-check, message=F, echo=F}

if (sum(duplicated(ss.df$unique_sample_code)) == 0 ){
  sample_metadata = gendiver::read.sample_sheet_lab(SS_URL)
  rm(ss.df)
  
  print("All Unique Sample Codes are UNIQUE -> OK!")

} else {
  
  print("Non Unique Sample Codes:")
  
  nonUSI = ss.df$unique_sample_code[duplicated(ss.df$unique_sample_code)]
  print(as.data.frame(table(ss.df$unique_sample_code[ss.df$unique_sample_code %in% nonUSI])))
  
  knitr::knit_exit()
}

```

## CHECK: USI codes are in expected format

- Unique Sample codes should be structured like in: [Guidelines Samplesheet voor een metabarcoding run](https://docs.google.com/document/d/1s3cuuwSkjyhIDy8Pd9Hf9wfjnXdV71aS)

```{r USI-complete, echo=F}

# unique_sample_code_regex = "\\d+_\\d+_\\d+_\\d+_\\d+_\\d+_\\d+_\\d+"
unique_sample_code_regex = "\\d{3}_\\d{3}_\\d{6}_\\d{3}_\\d{4}_\\d{3}_\\d{2}_\\d{5}"

print(paste0('Checking USI for format: ', unique_sample_code_regex))

if (sum(!grepl(unique_sample_code_regex, sample_metadata$UNIQUE_SAMPLE_CODE)) > 0){
  print(paste0("Warning: ", sum(!grepl(unique_sample_code_regex, sample_metadata$UNIQUE_SAMPLE_CODE)), " USIs are NOT in correct format!"))
  kable(sample_metadata$UNIQUE_SAMPLE_CODE[!grepl(unique_sample_code_regex, sample_metadata$UNIQUE_SAMPLE_CODE)], col.names = c("Incorrect USI format detected:"))
} else {
  
  print(paste0("All USIs conform with expected format -> OK !"))
  
}

```

## INFO: Overview of sample extracts and lab procedure

```{r labwork-overview, echo=F}

sample_metadata$TECHNICAL_REPLICATE_COLLAPSED = sample_metadata$TECHNICAL_REPLICATE
sample_metadata$TECHNICAL_REPLICATE_COLLAPSED[sample_metadata$TECHNICAL_REPLICATE_COLLAPSED > 3] = "3+"

kable(table(sample_metadata$PROJECT,
            sample_metadata$PRIMERS,
            sample_metadata$ANALYSE_TYPE,
            sample_metadata$TECHNICAL_REPLICATE_COLLAPSED), col.names = c("Project", "Primers","analyse", "replicate","PCR NR"))


```


## INFO: per POOL content

```{r pool-tables, echo=F, fig.pos='H'}
# XXX per pool
make.combi_table = function(sample_metadata, group=1, view=c("MATRIX", "PROJECT", "ANALYSE_TYPE","PRIMERS")) {
  
  tlist = as.list(1:length(view))
  tcols = as.list(1:length(view))
  
  for (i in tlist) {
    tlist[[i]] = table(sample_metadata[,group], sample_metadata[,view[i]])
    tcols[[i]] = paste0(colnames(sample_metadata)[view[i]], "=" ,colnames(tlist[[i]]))
  }
  
  combi_table = do.call(cbind, tlist)
  
  # colnames(combi_table) = unlist(tcols)
  
  return(combi_table)
}

tt = make.combi_table(sample_metadata)
kable(tt)
```


### Technical replicates

```{r pool_rep-table, echo=F}

kable(table(sample_metadata$LIBRARY_CODE, sample_metadata$TECHNICAL_REPLICATE))

```

## CHECK: barcode sequences are unique per pool

```{r bc-check, echo=F}

# Check unique barcodes per POOL
df1 = as.data.frame(table(sample_metadata$LIBRARY_CODE, sample_metadata$BARCODE_F_PRIMER))
df2 = as.data.frame(table(sample_metadata$LIBRARY_CODE, sample_metadata$BARCODE_R_PRIMER))
df1$TYPE="FWD"
df2$TYPE="REV"

bc.df = rbind(df1, df2)

BARCODE_CHECK=FALSE
if(max(bc.df$Freq) > 1){
  problem.bc = bc.df[bc.df$Freq > 1,]
  print("Warning: POOLs with NON-unique barcode sequences detected!")
  colnames(problem.bc) = c("LIBRARY_CODE", "BARCODE", "Freq", "TYPE")

  print(table(as.character(problem.bc$LIBRARY_CODE)))

  kable(problem.bc[, c(1,2,4,3)], row.names = FALSE)
} else {
    
  print(paste0("All barcodes UNIQUE within a POOL -> OK !"))
  BARCODE_CHECK=TRUE


  }

```

## CHECK: PLATE_NOTES should reflect physical/real plate

- Ideally (for downstream data analysis) PLATE_NOTES reflects a real plate -> **96 wells max**
- We need to define a "physical/real" PCR-plate as `PLATE_NOTES` + `PRIMER`

```{r PLATE_NOTES-check, echo=F}

sample_metadata$PLATE = paste0(sample_metadata$PRIMERS, ":", sample_metadata$PLATE_NOTES)
sample_metadata$SAMPLE_TYPE = paste0(sample_metadata$PROJECT,":", sample_metadata$PRIMERS)
sample_metadata$SAMPLE_TYPE[sample_metadata$ANALYSE_TYPE_CODE < 999 ] = sample_metadata$ANALYSE_TYPE[sample_metadata$ANALYSE_TYPE_CODE < 999]

plate_tab = table(sample_metadata$PLATE, sample_metadata$SAMPLE_TYPE)

kable(plate_tab)

if ( sum(rowSums(plate_tab) > 96) > 1 ){
  print("Warning: Some PLATE_NOTES contain more than 96 data-points")
  tt = table(sample_metadata$PLATE_NOTES)
  kable(tt[tt>96], col.names = c("PLATE_NOTES", "Number of samples (>96)"))

} else {
  print("PLATE_NOTES seems to reflect real plate (max 96 wells) -> OK!")
}

plh=max(3, length(unique(sample_metadata$PLATE_NOTES)))
```
\break

## INFO: Plate layout

```{r plate-layout, echo=F, fig.width=15, fig.height=plh}
sample_metadata$WELL_COLUMN = as.factor(sample_metadata$WELL_COLUMN)
sample_metadata$LABEL = sample_metadata$ANALYSE_TYPE
sample_metadata$LABEL[sample_metadata$ANALYSE_TYPE_CODE == 999] = ""
sample_metadata$LABEL = paste0(sample_metadata$LABEL, sample_metadata$TECHNICAL_REPLICATE_CODE)
mycols = length(unique(sample_metadata$PRIMERS))

ggplot2::ggplot( data=sample_metadata, ggplot2::aes(
  x = .data$WELL_COLUMN, y=.data$WELL_ROW, fill=.data$SAMPLE_TYPE)) +
ggplot2::geom_tile(colour = "black", linewidth=0.2) +
  ggplot2::geom_text(ggplot2::aes(label = LABEL), colour="black", size=2, alpha=0.8) +
ggplot2::facet_wrap(~.data$PLATE, ncol = 3, scales = "free") +
ggplot2::scale_x_discrete(limits=factor(c(1:12))) +
ggplot2::scale_y_discrete(limits=rev) +
ggplot2::theme_classic()
  # ggplot2::theme(legend.position="none")

```

\pagebreak

## Write barcode-files for demultiplexing

- Once the sample-sheet is evaluated and barcodes are OK -> make tsv files to use as input for demultiplexing

```{r make-barcode-files, echo=T}

if (BARCODE_CHECK==TRUE){
  print("INFO: Writing barcode files")
  outdir=params$BC_OUT_DIR
  dir.create(outdir)
  gendiver::write.barcode_files(sample_metadata, out_dir = outdir, LIB_COL = "LIBRARY_CODE")
}

```

