---
title: "Demultiplex Report"
author: 'Nick Dillen'
date: "`r Sys.Date()`"
output:
  pdf_document: 
    toc: true
    fig_caption: true
  html_document:
    code_folding: hide
    highlight: haddock
    number_sections: false
    fig_caption: true
    theme: sandstone
    toc: true
    toc_float: true
urlcolor: blue
params: 
  SAMPLE_SHEET_URL: "URL-sample-sheet"
  DEMULTIPLEX_RESULT_PATH: "PATH-to-02_demultiplex"
  OUT_DIR: "output-dir-supplements"
---

```{r read-param, echo=FALSE, eval=T}
SS_URL=params$SAMPLE_SHEET_URL
data_dir = params$DEMULTIPLEX_RESULT_PATH

print(params)
```


```{r prep, echo=F, message=F}
library(knitr)
library(gendiver)
library(stringr)
library(ggplot2)

googlesheets4::gs4_auth(T)

# -----------------------------------------------------
sample_sheet = read.sample_sheet_lab(SS_URL)
PROJ_NAME = basename(dirname(data_dir))

```

\break

# `r unique(sample_sheet$RUN)`

# Demultiplexed reads QC

```{r read.demultiplexed, echo=F}
read_count_track.df = read.summary_files_demultiplex(data_dir)
# read_count_track.df = read.summary_files_demultiplex(data_dir, summary_file_pattern = "_sample.txt", legacy_mode = T)

# only keep R1 data
read_count_track.df = read_count_track.df[grepl("1$|FWD$", read_count_track.df$type),]

# add total raw reads:
# because demultiplexing is done in 2 steps, unk+res is 2x raw
tot_reads_multiplexed = sum(read_count_track.df$read_count) / 2

```

## Demultiplexing success TOTAL

- Demultiplexing success complete run-level

```{r echo=F}
total_demultiplexed=sum(read_count_track.df$read_count[! read_count_track.df$sample_id == "unknown"])
demultiplex_success_percent = round(total_demultiplexed/(tot_reads_multiplexed)*100, digits=2)
```

- Demultiplexing success for the complete run is:  `r demultiplex_success_percent`%
- This corresponds to `r total_demultiplexed` demultiplexed reads out of `r tot_reads_multiplexed` total multiplexed reads


## Demultiplexing success per POOL

- Demultiplexing success POOL

```{r demultiplex-per-pool, fig.height=6, fig.width=16, echo=F}

demultiplex_summary_df_pp = aggregate(read_count ~ type + multiplexed_source, data = read_count_track.df, sum)

# add total raw reads per plate:
# because demultiplexing is done in 2 steps, unk+res is 2x raw
tot_raw_sum_pp = aggregate(read_count ~ multiplexed_source, data = demultiplex_summary_df_pp, sum)
tot_raw_sum_pp$read_count = tot_raw_sum_pp$read_count/2
tot_raw_sum_pp$type = "raw_data_R1"
demultiplex_summary_df_pp = rbind(demultiplex_summary_df_pp, tot_raw_sum_pp)

# create use info column
demultiplex_summary_df_pp$type_use = "used_data"
demultiplex_summary_df_pp$type_use[grepl(pattern = "unknown|discard|unassembled|5prime|Undetermined", x = demultiplex_summary_df_pp$type )] = "discarted_data"

# per plate demultiplexing success
pp_demu= demultiplex_summary_df_pp[demultiplex_summary_df_pp$type == "demultiplexed_R1",c("multiplexed_source", "read_count")]
pp_tot_raw = demultiplex_summary_df_pp[demultiplex_summary_df_pp$type == "raw_data_R1",c("multiplexed_source", "read_count")]

demultiplex_success_pp = merge(pp_demu, pp_tot_raw, by = 1)
demultiplex_success_pp$demultiplex_success_percent = demultiplex_success_pp$read_count.x/demultiplex_success_pp$read_count.y *100

demultiplex_summary_df_pp$type = factor(
  demultiplex_summary_df_pp$type, levels=c('raw_data_R1',"unknown_R1","unknown-round1-1","unknown-round1bis-1",'unknown_round2_1',"demultiplexed_R1"))

p_overview_data_pp = demultiplex_summary_df_pp[demultiplex_summary_df_pp$type_use == "used_data",]
p_overview_data_pp = merge(p_overview_data_pp, demultiplex_success_pp[, c(1,4)], by = "multiplexed_source")
p_overview_data_pp$demultiplex_success_percent[p_overview_data_pp$type == "raw_data_R1"] = 100

p_demult_overview_pp = ggplot(data=p_overview_data_pp, aes(x=type, y=read_count, fill=type)) +
  geom_col() + facet_grid(~multiplexed_source) +
  geom_text(
    aes(label = paste0(round(demultiplex_success_percent, 2), "%") ),
    position = position_dodge(width = 0.5), 
    vjust = -1,
  ) + 
  scale_x_discrete(labels = NULL, breaks = NULL) + labs(x = NULL) +
  ggtitle(paste0(PROJ_NAME, " - Overview of demultiplexing per pool"), subtitle = paste0("Combined demultiplexing success = ", demultiplex_success_percent, "%"))

p_demult_overview_pp

```


## Demultiplexing success per SAMPLE

Plotting the demultiplexed read_counts per sample, per POOL

- Data is ordered on POOL and then on read_count
- Empty samples/amplicon are in front per POOL

```{r echo=FALSE}
sfigh = min( 150, nrow(read_count_track.df)*0.2)
rc.df = read_count_track.df[read_count_track.df$type %in% c("demultiplexed_R1", "unknown_round2_1"),]
rownames(rc.df) = gsub("_R1.fastq.gz" , "", rc.df$file_name)
rc.df=rc.df[,"read_count", drop=F]
sample_sheet=merge(sample_sheet, rc.df, by = "row.names")
row.names(sample_sheet) = sample_sheet$Row.names
sample_sheet = sample_sheet[-1]
```


```{r p_demult_per_sample, fig.height=sfigh, fig.width=25, echo=F}

sample_order_filtered=sample_sheet$UNIQUE_SAMPLE_CODE[order(sample_sheet$LIBRARY_CODE, sample_sheet$read_count)]
p_summary = summary(sample_sheet$read_count)

p_demult_per_sample = ggplot(data=sample_sheet, aes(x=UNIQUE_SAMPLE_CODE, y=read_count, fill=LIBRARY_CODE)) +
  geom_bar(stat="identity", position = "dodge") + coord_flip() + 
  geom_hline(yintercept = p_summary) +
  annotate("text", x=length(unique(sample_sheet$UNIQUE_SAMPLE_CODE))-1, y=p_summary, angle = 90, label=names(p_summary)) +
  ggtitle(paste0(PROJ_NAME, " - Demultiplexed success per sample"),
          subtitle = paste(round(p_summary), names(p_summary), sep = "  :    ", collapse = "\n")) +
  scale_x_discrete(sample_order_filtered, limits = rev(sample_order_filtered), , name="UNIQUE_SAMPLE_CODE")
p_demult_per_sample
```


- This might be quite small for the number of samples/amplicons
- Check the generated supplement `.pdf`


```{r ggsave, echo=F}
ggfname=paste0("demultiplex-QC-report-supplement-fig.pdf")
ggoutp = file.path(params$OUT_DIR,ggfname)
print(ggfname)
ggsave(ggoutp, p_demult_per_sample, width = 25, height = sfigh, limitsize = FALSE)
```


## Demultiplexing success per PLATE (heatmap)

```{r plate-prep, echo=F}
sample_sheet$PLATE = paste0(sample_sheet$PRIMERS, ":", sample_sheet$PLATE_NOTES)
sample_sheet$SAMPLE_TYPE = paste0(sample_sheet$PROJECT,":", sample_sheet$PRIMERS)
sample_sheet$SAMPLE_TYPE[sample_sheet$ANALYSE_TYPE_CODE < 999 ] = sample_sheet$ANALYSE_TYPE[sample_sheet$ANALYSE_TYPE_CODE < 999]
# sample_sheet$SAMPLE_TYPE[sample_sheet$read_count < 250] = NA
sample_sheet$SAMPLE_TYPE = factor(sample_sheet$SAMPLE_TYPE)
sample_sheet$LABEL = sample_sheet$ANALYSE_TYPE
sample_sheet$LABEL[sample_sheet$ANALYSE_TYPE_CODE == 999] = ""
sample_sheet$WELL_COLUMN = as.factor(sample_sheet$WELL_COLUMN)
mycols = length(unique(sample_sheet$PRIMERS))

# some extra for next chunk
plh=length(unique(sample_sheet$PLATE_NOTES)) *1.5
```


- Map readcounts per WELL to **check overall read distribution**

```{r plate-layout-ident, echo=F, fig.width=15, fig.height=plh, warning=FALSE}
pl_plot = ggplot2::ggplot( data=sample_sheet, ggplot2::aes(
  x = .data$WELL_COLUMN, y=.data$WELL_ROW, fill=.data$SAMPLE_TYPE, alpha=.data$read_count, colour=.data$SAMPLE_TYPE)) +
ggplot2::geom_tile(linewidth=0.2, colour="black") +
  geom_text(aes(label = LABEL), colour="black", size=2, alpha=0.8) +
ggplot2::facet_wrap(~.data$PLATE, ncol = 3, scales = "free") +
ggplot2::scale_x_discrete(limits=factor(c(1:12))) +
ggplot2::scale_y_discrete(limits=rev) +
  scale_fill_discrete() + ggplot2::theme_classic()
pl_plot
```


- Map **log10** readcounts per WELL to **highlight samples/amplicon with low read-counts**

```{r plate-layout-log, echo=F, fig.width=15, fig.height=plh, warning=FALSE}
pl_plot + scale_alpha_continuous(transform="log10")
```

## Readcounts per CONTROL

```{r echo = F}
type.tb = table(sample_sheet$ANALYSE_TYPE)
type.tb

controls = names(type.tb)[! grepl("Metabarcoding|eDNA",names(type.tb))]

control_plots = as.list(controls, controls)
names(control_plots) = controls

for (typei in controls){
  
  cdata = sample_sheet[sample_sheet$ANALYSE_TYPE == typei,]
  
  p_summary = summary(cdata$read_count)
  sample_order_filtered=cdata$UNIQUE_SAMPLE_CODE[order(cdata$LIBRARY_CODE, cdata$read_count)]

  p_control = ggplot(data=cdata, aes(x=UNIQUE_SAMPLE_CODE, y=read_count)) +
    geom_col() + coord_flip() +
    ggtitle(paste0(PROJ_NAME, " - Demultiplexed reads: ",typei ),
            subtitle = paste(round(p_summary), names(p_summary), sep = "  :    ", collapse = "\n")) +
    scale_x_discrete(sample_order_filtered, limits = rev(sample_order_filtered), , name="UNIQUE_SAMPLE_CODE") + 
    xlab("test")
  
  control_plots[[typei]] = p_control
  
}

```


```{r fig.width=10, echo=F}

print(control_plots)

```

## Readcounts per barcode combination Per POOL

Inspect if certain barcodes perform consistently better/worse (per POOL)

```{r, echo=F}

sample_sheet$BARCODE_F_PRIMER = factor(sample_sheet$BARCODE_F_PRIMER)
bc_pool_df = aggregate(data=sample_sheet, read_count ~ WELL_ROW + WELL_COLUMN + LIBRARY_CODE + BARCODE_F_PRIMER + SAMPLE_TYPE, sum)

bc_pool_total = aggregate(data=sample_sheet, read_count ~ LIBRARY_CODE, sum)

bc_pool_df = merge(bc_pool_df, bc_pool_total, by="LIBRARY_CODE", suffixes = c("", "_pool"))
bc_pool_df$percent = (bc_pool_df$read_count / bc_pool_df$read_count_pool )* 100
# bc_order = aggregate(data=bc_pool_df, read_count~BARCODE_F_PRIMER, sum)
# bc_order_x =  bc_order$BARCODE_F_PRIMER[order(bc_order$read_count)]


bc_order = bc_pool_df[bc_pool_df$LIBRARY_CODE == unique(bc_pool_df$LIBRARY_CODE)[1],]
bc_order_x =  bc_order$BARCODE_F_PRIMER[order(bc_order$percent)]

p_bc_pool = ggplot(data=bc_pool_df, aes(x=BARCODE_F_PRIMER, y=percent, fill=LIBRARY_CODE)) + facet_grid(~LIBRARY_CODE) + coord_flip()+
  geom_bar(stat="identity", position = "dodge") + 
  ggtitle(paste0(PROJ_NAME, " - Demultiplexed success per BARCODE")) +
  ylab("Percent of reads (%)")

fh = length(bc_order_x)*0.15
fw = length(unique(bc_pool_df$LIBRARY_CODE))*2
```


```{r fig.width=fw, fig.height=fh, echo=F}
p_bc_pool
```

## BARCODE heatmap [EXPERIMENTAL]

```{r, echo=F}
barcode_heatmap = function(df, FUNY="mean", correct_nrsamples=T){
  
  if (correct_nrsamples){
    # correct for numbe of samples per pool
    dd = aggregate(data=df, BARCODE_F_PRIMER ~ LIBRARY_CODE, function(x){length(unique(x))})
    colnames(dd) = c("LIBRARY_CODE", "xfactor"     )
    dd$xfactor = dd$xfactor / max(dd$xfactor)
    
    df = merge(df, dd)
    df$read_count = df$read_count * df$xfactor
  }

  
  bc_df = aggregate(data=df, read_count ~ WELL_ROW + WELL_COLUMN + BARCODE_F_PRIMER, FUN=FUNY)
  # Try to guess number of sets
  check_sets = bc_df$BARCODE_F_PRIMER[bc_df$WELL_ROW == "A" & bc_df$WELL_COLUMN == "1"]
  
  bc_df$BC_SET = "SET_0"
  
  for (bc_i in 1:length(check_sets)){
    bc_df$BC_SET[as.integer(bc_df$BARCODE_F_PRIMER) >= as.integer(check_sets[bc_i])] = paste0("SET_", bc_i)
  }
  
  # bc_df$BC_SET[as.integer(as.character(bc_df$BARCODE_F_PRIMER)) > 250] = "SET_2"
  
  
  bc_df$read_count[bc_df$read_count < 1] = NA
  
  pout = ggplot2::ggplot( data=bc_df, ggplot2::aes(
    x = .data$WELL_COLUMN, y=.data$WELL_ROW, fill=.data$read_count)) +
  ggplot2::geom_tile(linewidth=0.2, colour="black") +
    geom_text(aes(label = BARCODE_F_PRIMER), colour="black", size=2.5) +
  ggplot2::facet_wrap(~.data$BC_SET, ncol = 3, scales = "free") +
  ggplot2::scale_x_discrete(limits=factor(c(1:12))) +
  ggplot2::scale_y_discrete(limits=rev) +
    scale_fill_continuous(type="viridis",
      transform="identity",
      na.value="white") +
  ggplot2::theme_classic() + ggtitle(paste0("Read counts (",FUNY,") per barcode for all libraries combined")) +
  labs(fill = paste0(FUNY," read count")) 
  
  return(pout)
  
}
```

- Check if pools have the same number of pooled amplicons
- Depending on the summary statistic used this might bias the signal in the following plots!

```{r barcode-layout, fig.width=12, fig.height=4}
pt = table(sample_sheet$LIBRARY_CODE)

# print(paste0("Removing ", names(pt[pt!=2*96]) , " from Barcode analysis"))

kable(pt)

# ssdf=sample_sheet[sample_sheet$LIBRARY_CODE %in% names(pt[pt==2*96]),]
ssdf = sample_sheet

barcode_heatmap(ssdf)
barcode_heatmap(ssdf, "median")
barcode_heatmap(ssdf, "min" )
barcode_heatmap(ssdf, "max")

```


## SUPPLEMENT: Readcount-table for detailed investigation

- If there are more run/data specific questions regarding the demultiplexing success please find the supplementary table and investigate as you like!

```{r echo=FALSE}
tfname=paste0("demultiplex-QC-report-supplement-table.csv")
tfoutp = file.path(params$OUT_DIR, tfname)
print(tfname)
sample_sheet$LABEL=NULL
write.table(sample_sheet, file = tfoutp, sep=",", row.names = F)
```
- You can read this in R with:

```{r, eval=F}
mydf = read.csv("demultiplex-QC-report-supplement-table.csv", header=T)
```
