---
title: "Pre-processing Report"
author: 'Nick Dillen'
date: "`r Sys.Date()`"
output:
  pdf_document: 
    toc: true
    fig_caption: true
  html_document:
    code_folding: hide
    highlight: haddock
    number_sections: false
    fig_caption: true
    theme: sandstone
    toc: true
    toc_float: true
urlcolor: blue
params: 
  SAMPLE_SHEET_URL: "URL-sample-sheet"
  PREPROCESS_RESULT_PATH: "PATH-to-03_pre_process"
  OUT_DIR: "PATH-to-00_preprocessQC"
---

```{r read-param, echo=FALSE, eval=T}
SS_URL=params$SAMPLE_SHEET_URL
data_dir = params$PREPROCESS_RESULT_PATH

print(params)
```


```{r prep, echo=F, message=F}
library(knitr)
library(gendiver)
library(stringr)
library(ggplot2)

googlesheets4::gs4_auth(T)

# -----------------------------------------------------
sample_sheet = read.sample_sheet_lab(SS_URL)
PROJ_NAME = paste0(basename(dirname(data_dir)), ":", basename(data_dir))
```

\break

# `r unique(sample_sheet$RUN)`
# `r PROJ_NAME`

# Pre-process reads QC

Input used:

```{r read.preprocessed, echo=F}
PPP_dirs = list.dirs(data_dir, recursive = T) 

PPP1 = tail(PPP_dirs[grepl("1_merged", PPP_dirs)], n=1)
PPP2 = tail(PPP_dirs[grepl("2_trimmed", PPP_dirs)], n=1)
PPP3 = tail(PPP_dirs[grepl("3_quality", PPP_dirs)], n=1)

# VERBOSE: print dirs
cat(data_dir, ":")
cat(paste(sapply(c(PPP1, PPP2, PPP3),basename), collapse = "\n"))

prepro_df = dplyr::bind_rows(
  read.logs_preprocess(PPP1, stepname = "merged"),
  read.logs_preprocess(PPP2, stepname = "trimmed", logpattern = "trimmed.log"),
  read.logs_preprocess(PPP3, stepname = "filter", logpattern = "filter.log")
)

prepro_df$pipeline_step = "PRE-PROCESS"

```

- calculate original per-sample DEMULTIPLEXED input from merging logs/files

```{r add-demu, echo=F}

if ("vsearch-merging" %in% prepro_df$toolname){
  print('vsearch-merging detected!')
  per_sample_demultiplexed_input = data.frame(
    row.names = NULL,
    "read_count"=sapply(list.files(PPP1, pattern = ".log", full.names = T), extract.mergepairs_log.tot_input),
    "sample_id"=gsub("\\..*", "",list.files(PPP1, pattern = ".log")),
    "file_name"=NA,
    "pipeline_step" = "DEMULTIPLEX",
    "type" = "total-demultiplexed",
    "toolname" = "demultiplexing"
    )
  

} else if ("pear-merging" %in% prepro_df$toolname){
  print('pear-merging detected!')
  per_sample_demultiplexed_input = aggregate(
    read_count ~ sample_id ,
    data = prepro_df[prepro_df$type %in% c("discarded", "assembled", "unassembled.forward"),],
    sum)

  per_sample_demultiplexed_input$file_name = NA
  per_sample_demultiplexed_input$type = "total-demultiplexed"
  per_sample_demultiplexed_input$pipeline_step = "DEMULTIPLEX"
  per_sample_demultiplexed_input$toolname = "demultiplexing"

} else {
  print('NO merge/assembly info !')
}


total_demultiplexed_input = sum(per_sample_demultiplexed_input$read_count)

prepro_df = rbind(per_sample_demultiplexed_input, prepro_df)

# Order the types following pre-processing pipeline
process_order = c(
  "total-demultiplexed", # demultiplexing
  "unassembled.forward", "unassembled.reverse", "discarded", "assembled", # PEAR output
  "merged", "mergepairs", # vsearch --fastq_mergepairs output
  "trimmed-5prime", "fullytrimmed", #cutadapt output
  "filtered", "fastqfilter" # vsearch fastq_filter output
)
prepro_df$type = factor(prepro_df$type, levels = rev(process_order), ordered = TRUE)
prepro_df$toolname = factor(prepro_df$toolname, levels=(unique(prepro_df$toolname)), ordered=TRUE)

```

## Pre-processing success TOTAL

- Pre-processing success run-level (amplicon specific)

```{r calc-total, echo=F}
# Calculate success %
preprocess_total_reads=sum(prepro_df$read_count[grepl("filter", prepro_df$type) ])
preprocess_success_percent = round(preprocess_total_reads / total_demultiplexed_input *100, 2)

# Calculate overview read numbers
ovsum = aggregate(prepro_df$read_count, by=list(prepro_df$type), FUN=sum)
ovsum_txt = paste("  - ", rev(paste(ovsum$Group.1,ovsum$x, sep = ": ")), collapse = "\n")
```

- Pre-processing success is:  `r preprocess_success_percent`%
- This corresponds to `r preprocess_total_reads` reads passing pre-processing out of `r total_demultiplexed_input` total demultiplexed reads


```{r pp-overview-plot, fig.height=5, fig.width=10, echo=F}
## TOTAL READCOUNT PER STEP
p_overview <- ggplot(data=prepro_df, aes(x=type, y=read_count, fill=toolname)) +
  geom_col() + coord_flip() + theme_minimal() +
  xlab("File type") +
  ylab("Total number of reads") +
  ggtitle(paste0(PROJ_NAME),
          subtitle = paste0(
            paste0("Pre-processing success (filtered/demultiplexed) = ", preprocess_success_percent, "%"),
            "\n", ovsum_txt))

p_overview
```


## Pre-processing success per POOL

```{r pp-success, echo=F}
for (stepi in levels(prepro_df$toolname)){
  tempdf = prepro_df[prepro_df$toolname == stepi,c("read_count", "sample_id")]
  colnames(tempdf) = c(paste0("read_count.", stepi),"UNIQUE_SAMPLE_CODE")
  sample_sheet = merge(sample_sheet, tempdf)
}
read_count.preprocess = tail(colnames(sample_sheet), 1)

sample_sheet$percent.preprocess = round((sample_sheet[[read_count.preprocess]] / sample_sheet$read_count.demultiplex) * 100, 2)
```


```{r pp-per-pool, fig.height=6, fig.width=16, echo=F}

pp_summary_df_pp = aggregate(
  cbind(sample_sheet$read_count.demultiplex, sample_sheet[[read_count.preprocess]]) ~  sample_sheet$LIBRARY_CODE,FUN =  sum)
colnames(pp_summary_df_pp) = c("LIBRARY_CODE", "read_count.demultiplex", "read_count.preprocess")
pp_summary_p = reshape2::melt(pp_summary_df_pp, id=1, value.name = "read_count", )
pp_summary_p$variable = as.character(pp_summary_p$variable)

pp_summary_p = merge(pp_summary_p, pp_summary_df_pp[, c("LIBRARY_CODE", "read_count.demultiplex"), drop=F], by="LIBRARY_CODE" )

pp_summary_p$percent.preprocess = round((pp_summary_p$read_count / pp_summary_p$read_count.demultiplex) * 100,2)

p_pp_pp = ggplot(data=pp_summary_p, aes(x=variable, y=read_count, fill=variable)) +
  geom_col() + facet_grid(~LIBRARY_CODE) +
  geom_text(
    aes(label = paste0(round(percent.preprocess, 2), "%") ),
    position = position_dodge(width = 0.5), 
    vjust = -1,
  ) + 
  scale_x_discrete(labels = NULL, breaks = NULL) + labs(x = NULL) +
  ggtitle(paste0(PROJ_NAME, " - Overview of pre-processing per pool"))

p_pp_pp

```


## Pre-processing success per SAMPLE/AMPLICON

Plotting the pre-processed read_counts per sample/amplicon

- Summary statistics on all samples/amplicons:

```{r sum-sts, echo=F}
filter_counts = prepro_df$read_count[grepl("filter", prepro_df$type)]
summary(filter_counts)
```

- Quantiles in more detail:

```{r qtiles, echo=F}
kable(floor(quantile(filter_counts, seq(0.05,0.95,0.05))), col.names = c("Quantile", "Readcounts"))
```

- Per SAMPLE/AMPLICON number of reads, grouped per POOL:

```{r prep-plot, echo=F}
prepro_df = merge(prepro_df, sample_sheet[, c("UNIQUE_SAMPLE_CODE", "LIBRARY_CODE")], by.x="sample_id", by.y="UNIQUE_SAMPLE_CODE")

sfigh = min(150, max(table(sample_sheet$LIBRARY_CODE)) * 0.5)

qcplot.preprocessing_sample = function(prepro_df){
  ## PER SAMPLE READCOUNT tracking
sample_read_count_track.df = prepro_df[prepro_df$sample_id != "no_sample",]
# And add total reads per sample
per_sample_total = sample_read_count_track.df[sample_read_count_track.df$type=="total-demultiplexed", c("sample_id", "read_count")]
colnames(per_sample_total) = c("sample_id", "read_count_total")
sample_read_count_track.df = merge(sample_read_count_track.df, per_sample_total, by='sample_id')
# Calc percentage
sample_read_count_track.df$read_count_percent = sample_read_count_track.df$read_count / sample_read_count_track.df$read_count_total * 100

# Prune data
discarded_types = levels(sample_read_count_track.df$type)[grepl("discard|unassembled|REV|unknown", levels(sample_read_count_track.df$type))]
sample_read_count_track.df = sample_read_count_track.df[!sample_read_count_track.df$type %in% discarded_types,]

# Order dataframe on type -> nice overlap order to track readcounts per step
sample_read_count_track.df = sample_read_count_track.df[order(sample_read_count_track.df$type, decreasing = T),]

# order sample-id levels on filtered read_counts -> order samples/amplicon based on RESULTS (in plot)
per_sample_total_filtered = sample_read_count_track.df[grepl("filter", sample_read_count_track.df$type), c("sample_id", "read_count", "read_count_percent")]
sample_order_filtered = per_sample_total_filtered$sample_id[order(per_sample_total_filtered$read_count)]
sample_order_filtered_percent = per_sample_total_filtered$sample_id[order(per_sample_total_filtered$read_count_percent)]

# Calculate some summary statistics for the final product
r_sum = summary(sample_read_count_track.df$read_count[grepl("filter", sample_read_count_track.df$type)])
p_sum = summary(sample_read_count_track.df$read_count_percent[grepl("filter", sample_read_count_track.df$type)])

# Plot total reads
p_sample_overview <- ggplot(data=sample_read_count_track.df, aes(x=sample_id, y=read_count, fill=type)) +
  geom_bar(stat="identity", position = position_identity()) + coord_flip() +
  xlim(sample_order_filtered) + # "hack" to create custom order on x axis
  scale_fill_brewer(palette = "Accent") +
  geom_hline(yintercept = r_sum) + 
  annotate("text", x=length(unique(sample_read_count_track.df$sample_id))-1, y=r_sum, angle = 90, label=names(r_sum)) +
  xlab("Unique sample code") +
  ylab("Number of reads\n(horizontal lines correspond to summary statistics)") +
  ggtitle(paste0(PROJ_NAME, " - Per sample read-counts"),
          subtitle = paste(round(r_sum), names(r_sum), sep = "  :    ", collapse = "\n"))

# plot percentage of demultiplexed
p_sample_overview_percent = ggplot(
  data=sample_read_count_track.df, aes(x=sample_id, y=read_count_percent, fill=type)) +
  geom_bar(stat="identity", position = position_identity()) + coord_flip() + 
  xlim(sample_order_filtered_percent) + # "hack" to create custom order on x axis
  scale_fill_brewer(palette = "Accent") +
  xlab("Unique sample code") +
  ylab("Percentage of reads (%)") +
  ggtitle(paste0(PROJ_NAME, " - Per sample read-count-percent (%)"),
          subtitle = paste(round(p_sum,2), names(p_sum), sep = "  :    ", collapse = "\n"))

return(list(p_sample_overview, p_sample_overview_percent))
}
```

```{r per-pool-pp-plot, fig.height=sfigh, fig.width=20, echo=F, results='asis'}

for (POOL in unique(prepro_df$LIBRARY_CODE)){
  pool_p = qcplot.preprocessing_sample(prepro_df[prepro_df$LIBRARY_CODE == POOL,])
  outp = pool_p[[1]]
  print(outp + ggplot2::ggtitle(paste0(POOL, " - ", outp$labels$title), subtitle = outp$labels$subtitle))
  # print(pool_p[[2]])
}

```

\break

## Pre-processing success per PLATE (heatmap)

- Using the readcounts per amplicon after pre-processing (`filter`)

```{r prep-hm, echo=FALSE}
sample_sheet$PLATE = paste0(sample_sheet$PRIMERS, ":", sample_sheet$PLATE_NOTES)
sample_sheet$SAMPLE_TYPE = paste0(sample_sheet$PROJECT,":", sample_sheet$PRIMERS)
sample_sheet$SAMPLE_TYPE[sample_sheet$ANALYSE_TYPE_CODE < 999 ] = sample_sheet$ANALYSE_TYPE[sample_sheet$ANALYSE_TYPE_CODE < 999]
# sample_sheet$SAMPLE_TYPE[sample_sheet$read_count < 250] = NA
sample_sheet$SAMPLE_TYPE = factor(sample_sheet$SAMPLE_TYPE)
sample_sheet$LABEL = sample_sheet$ANALYSE_TYPE
sample_sheet$LABEL[sample_sheet$ANALYSE_TYPE_CODE == 999] = ""
sample_sheet$WELL_COLUMN = as.factor(sample_sheet$WELL_COLUMN)
mycols = length(unique(sample_sheet$PRIMERS))

figh = length(unique(sample_sheet$PLATE_NOTES)) * 4
```

- Map readcounts per WELL to **check overall read distribution**

```{r plate-layout-ident, echo=F, fig.width=15, fig.height=figh, warning=FALSE}

pl_p = ggplot2::ggplot( data=sample_sheet, ggplot2::aes(
  x = .data$WELL_COLUMN, y=.data$WELL_ROW, fill=.data$SAMPLE_TYPE, alpha=.data[[read_count.preprocess]], colour=.data$SAMPLE_TYPE)) +
ggplot2::geom_tile(linewidth=0.2, colour="black") +
  geom_text(aes(label = LABEL), colour="black", size=2, alpha=0.8) +
ggplot2::facet_wrap(~.data$PLATE, ncol = 3, scales = "free") +
ggplot2::scale_x_discrete(limits=factor(c(1:12))) +
ggplot2::scale_y_discrete(limits=rev) +
  scale_fill_discrete() +
ggplot2::theme_classic()

pl_p

```

- Map **log10** readcounts per WELL to **highlight samples/amplicons with low read-counts**


```{r plate-layout-log, echo=F, fig.width=15, fig.height=figh, warning=FALSE}

pl_p + scale_alpha_continuous(transform="log10")

```

## Readcounts per CONTROL

```{r prep-c, echo = F}
type.tb = table(sample_sheet$ANALYSE_TYPE)
type.tb

controls = names(type.tb)[! grepl("Metabarcoding|eDNA",names(type.tb))]

control_plots = as.list(controls, controls)
names(control_plots) = controls

fH = min(150, max(type.tb[controls])*0.2)
```



```{r plot-c, fig.height=fH, fig.width=20, echo=F, results='asis'}

for (typei in controls){
  cdata_samples = sample_sheet$UNIQUE_SAMPLE_CODE[sample_sheet$ANALYSE_TYPE == typei]
  control_p = qcplot.preprocessing_sample(prepro_df[prepro_df$sample_id %in% cdata_samples,])
  outp = control_p[[1]]
  print(outp + ggplot2::ggtitle(paste0(typei, " - ", outp$labels$title), subtitle = outp$labels$subtitle))
  # print(pool_p[[2]])
}

```

## SUPPLEMENT: Readcount-table for detailed investigation

- If there are more run/data specific questions regarding the pre-processing, please find the supplementary table and investigate as you like!

```{r supp-tab, echo=FALSE}
tfname=paste0("preprocess-QC-report-supplement-table.csv")
tfoutp = file.path(params$OUT_DIR, tfname)
print(tfname)
sample_sheet$LABEL=NULL
write.table(sample_sheet, file = tfoutp, sep=",", row.names = F)
# If this throws an error: in `write.table()`:! unimplemented type 'list' in 'EncodeElement'
# Need to change the extraction date in the sample sheet format -> numeric 000000 (also update USI col probably)
```
- You can read this in R with:

```{r show-ex, eval=F}
mydf = read.csv("preprocess-QC-report-supplement-table.csv", header=T)
```
