---
title: "Basic analysis Report"
author: 'Nick Dillen'
date: "`r Sys.Date()`"
output:
  pdf_document: 
    toc: true
    fig_caption: true
  html_document:
    code_folding: hide
    highlight: haddock
    number_sections: false
    fig_caption: true
    theme: sandstone
    toc: true
    toc_float: true
urlcolor: blue
params: 
  SAMPLE_SHEET_URL: "URL-sample-sheet"
  TAX_PATH: "TAX-result"
  OTU_PATH: "OTU-table"
  FASTA_PATH: "FASTA"
  OUT_DIR: "."
---

# Operational Data report

- Investigate basic statistic of combined **Clustering** and **Taxonomy** for a certain dataset
- Interpret and evaluate **non-experiment** samples/amplicons

## Input data

- Input lab sample sheet: `r params$SAMPLE_SHEET_URL`
- Input OTU table: `r params$OTU_PATH`
- Input Taxonomy file: `r params$TAX_PATH`

```{r read-param, echo=FALSE, eval=T, message=F}
library(gendiver)
library(knitr)
library(ggplot2)
library(phyloseq)
library(Biostrings)

googlesheets4::gs4_auth(T)
googledrive::drive_auth(T)

otu_tab_path = params$OTU_PATH
tax_tab_path = params$TAX_PATH
fasta_path = params$FASTA_PATH

# Check file location
if (! file.exists(otu_tab_path)) {
  otu_tab_path = googledrive::drive_get(otu_tab_path)
  otu_tab_path = otu_tab_path$name
}

if (! file.exists(tax_tab_path)) {
  tax_tab_path = googledrive::drive_get(tax_tab_path)
  tax_tab_path = tax_tab_path$name
}

if (! file.exists(fasta_path)) {
  fasta_path = googledrive::drive_get(fasta_path)
  fasta_path = fasta_path$name
}

PROJ_NAME = gsub("\\..*", "", basename(otu_tab_path))
SS_NAME = googledrive::drive_get(params$SAMPLE_SHEET_URL)$name
```

- Used lab sample sheet: `r SS_NAME`
- Used OTU table: `r basename(otu_tab_path)`
- Used Taxonomy file: `r basename(tax_tab_path)`
- Used sequence file: `r basename(fasta_path)`


\break

# Read data

```{r read.opdata, echo=T}

sample_sheet = read.sample_sheet_lab(params$SAMPLE_SHEET_URL)

if (grepl("obitools", otu_tab_path, ignore.case = T)){
  otu_df <- read.otu_table_obitools3(params$OTU_PATH)
} else {
  #Read OTU table and tax table, and explore data
  otu_df <- read.otu_table_vsearch(params$OTU_PATH)
}

if (grepl("sintax", tax_tab_path, ignore.case = T)){
  tax_df =  read.taxonomy_sintax(params$TAX_PATH, clean_tax = T, add_custom = T)
} else if(grepl("idtaxa", tax_tab_path, ignore.case = T)){
  #Read OTU table and tax table, and explore data
  tax_df <- read.taxonomy_idtaxa(params$TAX_PATH)
} else if(grepl("ecotag", tax_tab_path, ignore.case = T)){
  #Read OTU table and tax table, and explore data
  tax_df <- read.taxonomy_obitools3(params$TAX_PATH)
} 

seqs = read.fastx_gdrive(params$FASTA_PATH)
```

# Basic exploration

- This is (for now) hardcoded for SINTAX and VSEARCH

```{r, fig.width=8, echo=F}

## BASIC EXPLORATION PLOTS
annot_overview=colSums(!is.na(tax_df))
annot_overview = annot_overview[names(annot_overview) != "custom_taxon"]
annot_overview = c(nrow(tax_df), annot_overview)
names(annot_overview)[1] = 'TOTAL'
print(annot_overview, row.names = F)

barplot(annot_overview, main="Taxonomy assignments overview", ylab = "Number of OTUs")

asv_lengths = width(seqs)

hist(asv_lengths, breaks = 100, main = "OTU seq length", sub=basename(otu_tab_path) ,xlab = "Length (bp)")


reads_length = data.frame("reads"=rowSums(otu_df),
                             "length"=asv_lengths, row.names = row.names(tax_df), width=12)
len_reads.df = aggregate(reads~length, reads_length, sum)

plot(len_reads.df, main = "Per ASV-length the number of reads",
     sub=basename(otu_tab_path), xlab = "ASV length")

plot(cbind(len_reads.df$length, cumsum(len_reads.df$reads)), main = "Per ASV-length the number of reads",
     sub=basename(otu_tab_path), xlab = "ASV length", ylab="Cumulative sum reads")

hist(rowSums(otu_df), breaks = 100, main = "Per taxa reads", sub=basename(otu_tab_path), 
     xlab = "Total number of reads")


hist(colSums(otu_df), breaks = 100, main = "Per sample reads", sub=basename(otu_tab_path),
     xlab = "Total number of reads")
```


# Make Phyloseq object

- To work with Phyloseq, check the documentation:
  - https://joey711.github.io/phyloseq/

```{r}
# Make phyloseq object
ps = phyloseq(
  otu_table(otu_df, taxa_are_rows = TRUE),
  tax_table(as.matrix(tax_df)),
  sample_data(sample_sheet),
  seqs
)

ps 
```

## Check CONTROL samples/amplicons

```{r}
type.tb = table(sample_data(ps)$ANALYSE_TYPE)
type.tb

controls = names(type.tb)[names(type.tb) != "Metabarcoding"]

control_plots = as.list(controls, controls)
names(control_plots) = controls

fH = max(type.tb[controls])*0.5
```

```{r, fig.width=20, fig.height=fH}
for (typei in controls){
  ps_control = subset_samples(ps, ANALYSE_TYPE==typei)
  ps_control = prune_taxa(rowSums(otu_table(ps_control)) > 1, ps_control)
  ps_control = tax_glom(ps_control, "species")
  ps_control
  
  # # create custom color
  txs = taxa_sums(ps_control)
  topx=20
  txs_keep = names(txs[order(txs, decreasing = T)][1:min(length(txs),topx)])
  txd = as.data.frame(tax_table(ps_control))
  txd$top_species = NA
  txd[txs_keep, "top_species"] = txd[txs_keep, "species"]

  tax_table(ps_control) = tax_table(as.matrix(txd))
  
  p_overview_control = plot_bar(ps_control, fill = "custom_taxon", x = "INDIVIDUAL") + 
  facet_grid(rows = vars(TECHNICAL_REPLICATE))+
  ggtitle(paste(typei, "- overview"), subtitle = PROJ_NAME)
  print(p_overview_control)
  
  p_overview_control_species = plot_bar(ps_control, fill = "top_species", x = "INDIVIDUAL") + 
  facet_grid(rows = vars(TECHNICAL_REPLICATE))+
  ggtitle(paste(typei, "- top",topx,"taxa"), subtitle = PROJ_NAME)
  print(p_overview_control_species)
}
```


## Overview

```{r MB-samples-overview}
### Inspect Samples ###
ps_samples = subset_samples(ps, ANALYSE_TYPE_CODE=="999")

# Look at top x ASVs (reduce for taxglom)
x_cutoff= 5000
topX = names(taxa_sums(ps_samples)[order(taxa_sums(ps_samples), decreasing = T)][1:min(ntaxa(ps_samples) ,x_cutoff)])

ps_topX = prune_taxa(topX, ps_samples)

ps_topX = tax_glom(ps_topX, "custom_taxon")
```

- Remove control samples
- Only look at top `r x_cutoff` top abundant ASVs (reads) to reduce data -> this amounts to `r round(sum(taxa_sums(ps_topX))/sum(taxa_sums(ps_samples)) *100, 2)`% of data
- group on custom taxonomy

```{r, fig.height=25, fig.width=30, warning=F}
normalized_ps = transform_sample_counts(ps_topX, function(OTU) OTU/sum(OTU) * 100)
  
plot_bar(normalized_ps, fill = "custom_taxon", x="INDIVIDUAL") + 
  # coord_flip() +
  facet_grid(rows = vars(TECHNICAL_REPLICATE)) + 
  ggtitle(paste(PROJ_NAME)) + ylab('Relative abundance (%)') +  theme(legend.position = "bottom")

```


# Taxonomy based clustering metrics

We can use taxonomy information to assess clustering.

- Note that taxonomy is often not real "Ground-Truth":
  - **Reference databases** have mistakes
  - Taxonomic annotations **tool** can make false positive/negative taxonomic annotations
  - **Noisy data** will lead to increased false positive annotations, picking up noise as signal. (=> data not enough filtered)
  - Target **amplicon** might not resolve into species, or have intra-species variation.


```{r}

ps_select = prune_taxa(topX, ps_samples)
pstab = as.data.frame(tax_table(ps_select))

taxdata = data.frame()
minprint = 15

pl_list = list()
tb_list= list()

for (tax_rank in colnames(pstab)){
  
  if (grepl("custom", tax_rank)){next}
  
  # cat(paste0("### ", tax_rank,"\n"))
  
  
  ddi = data.frame(table(pstab[[tax_rank]]))
  colnames(ddi) = c("Taxonomy", "nr_members")
  
  top10 = ddi[order(ddi$nr_members, decreasing = T)[1:min(minprint,nrow(ddi))],]
  
  if (nrow(ddi) < minprint){
    myc = "Taxa with number of members"
  } else {
    myc = paste("Top",minprint,"taxa with most members")
  }
  
  kbt = kable(top10, caption = myc, col.names = c(tax_rank, "nr_members"),
              row.names = FALSE, )
  
  tb_list[[tax_rank]] = kbt
  
  di = data.frame(table(table(pstab[[tax_rank]])))
  di$Var1 = as.numeric(as.character(di$Var1))
  colnames(di) = c("Members", "Taxa")
  
  if (nrow(di) > minprint && (sum(di$Taxa[1:5]) > sum(tail(di$Taxa)))){
    di$Members[di$Members > 100] = 101
    pl=   ggplot(data=di, aes(x = Members, y=Taxa)) + geom_col() + 
      ggtitle(tax_rank) + xlab("Members per Taxa (100+ is grouped)") +
      ylab("Taxa (freq)")
    
    pl_list[[tax_rank]] = pl
  } 
  
  
}


```


```{r}
tb_list
```



```{r}
pl_list
```

