---
title: "Operational Data Report"
author: 'Nick Dillen'
date: "`r Sys.Date()`"
output:
  # pdf_document: 
  #   toc: true
  #   fig_caption: true
  html_document:
    code_folding: hide
    highlight: haddock
    number_sections: false
    fig_caption: true
    theme: sandstone
    toc: true
    toc_float: true
urlcolor: blue
params:
  SAMPLE_SHEET_URL: "URL-sample-sheet"
  TAX_PATH: "TAX-result"
  OTU_PATH: "OTU-table"
  FASTA_PATH: "FASTA-path"
  OUT_DIR: "."
---

```{r setup, echo=FALSE,}
options(width = 80)
```


# Operational Data report

- Investigate basic statistic of combined **Clustering** and **Taxonomy** for a certain dataset
- Interpret and evaluate **non-experiment** samples/amplicons

## Input data

- Input lab sample sheet: `r params$SAMPLE_SHEET_URL`
- Input OTU table: `r params$OTU_PATH`
- Input Taxonomy file: `r params$TAX_PATH`
- Input FASTA: `r params$FASTA_PATH`

```{r read-param, echo=FALSE, eval=T, message=F}
library(gendiver)
library(knitr)
library(ggplot2)
library(phyloseq)
library(Biostrings)

googlesheets4::gs4_auth(T)
googledrive::drive_auth(T)

otu_tab_path = params$OTU_PATH
tax_tab_path = params$TAX_PATH
fasta_path = params$FASTA_PATH

# Check file location
if (! file.exists(otu_tab_path)) {
  otu_tab_path = googledrive::drive_get(otu_tab_path)
  otu_tab_path = otu_tab_path$name
}

if (! file.exists(tax_tab_path)) {
  tax_tab_path = googledrive::drive_get(tax_tab_path)
  tax_tab_path = tax_tab_path$name
}

if (! file.exists(fasta_path)) {
  fasta_path = googledrive::drive_get(fasta_path)
  fasta_path = fasta_path$name
}

PROJ_NAME = gsub("\\..*", "", basename(otu_tab_path))
SS_NAME = googledrive::drive_get(params$SAMPLE_SHEET_URL)$name
```

- Used lab sample sheet: `r SS_NAME`
- Used OTU table: `r basename(otu_tab_path)`
- Used Taxonomy file: `r basename(tax_tab_path)`
- Used sequence file: `r basename(fasta_path)`


\break

# Read data

```{r read.opdata, echo=F, warning=F, message=T}

sample_sheet = read.sample_sheet_lab(params$SAMPLE_SHEET_URL)

seqs = read.fastx_gdrive(params$FASTA_PATH)
names(seqs) = gsub(";size=\\d*", "", names(seqs))

if (grepl("obitools", otu_tab_path, ignore.case = T)){
  otu_df <- read.otu_table_obitools3(params$OTU_PATH)
} else {
  #Read OTU table and tax table, and explore data
  otu_df <- read.otu_table_vsearch(params$OTU_PATH)
}

if (grepl("sintax", tax_tab_path, ignore.case = T)){
  tax_df =  read.taxonomy_sintax(params$TAX_PATH, clean_tax = T, add_custom = T)
  TAX_TOOL="SINTAX"
} else if(grepl("idtaxa", tax_tab_path, ignore.case = T)){
  #Read OTU table and tax table, and explore data
  tax_df <- read.taxonomy_idtaxa(params$TAX_PATH)
  TAX_TOOL="IDTAXA"
} else if(grepl("obitools3", tax_tab_path, ignore.case = T)){
  #Read OTU table and tax table, and explore data
  tax_df_raw <- read.taxonomy_obitools3(params$TAX_PATH)
  tax_df = select.taxonomy_obitools3(tax_df_raw, min_BEST_ID = 1, fillNA = F)
  
  
  # seqs = seqs[row.names(tax_df)]
  TAX_TOOL="OBITOOLS3"

} 

```

\break

# Basic exploration of Operational Data

```{r basic-explore, fig.width=10, echo=F, fig.align='center'}

## BASIC EXPLORATION PLOTS
annot_overview=colSums(!is.na(tax_df))
annot_overview = annot_overview[names(annot_overview) != "custom_taxon"]
annot_overview = c(nrow(tax_df), annot_overview)
names(annot_overview)[1] = 'TOTAL'
full_sub =paste0(basename(otu_tab_path), "\n",basename(tax_tab_path))
barplot(annot_overview, main="Taxonomy assignments overview - OTUs", ylab = "Number of OTUs", sub=full_sub)
print(annot_overview, row.names = F)

tax_df_rc = tax_df
otu_tot_rc = rowSums(otu_df)
tax_df_rc$read_count = NA
tax_df_rc[names(otu_tot_rc), c("read_count")] = otu_tot_rc

# remove taxonomy not in OTU table
tax_df_rc = tax_df_rc[!is.na(tax_df_rc$read_count),]

tb_rc = sapply(colnames(tax_df), function(x) sum(tax_df_rc$read_count[!is.na(tax_df_rc[[x]])]))
annot_overview2 = tb_rc[names(tb_rc) != "custom_taxon"]
annot_overview2 = c(sum(tax_df_rc$read_count), annot_overview2)
names(annot_overview2)[1] = 'TOTAL'
barplot(annot_overview2, main="Taxonomy assignments overview - READS", ylab = "Number of reads", sub=full_sub)
print(annot_overview2, row.names = F)

# Sequence length
# remove sequences not in OTU table
seqs = seqs[rownames(otu_df)]
asv_lengths = width(seqs)

hist(asv_lengths, breaks = 100, main = "All data: OTU seq length", sub=basename(fasta_path) ,xlab = "Length (bp)")

reads_length = data.frame("reads"=rowSums(otu_df), "length"=asv_lengths)

len_reads.df = aggregate(reads~length, reads_length, sum)

plot(len_reads.df, main = "All data: Per ASV-length the number of reads",
     sub=basename(otu_tab_path), xlab = "ASV length")

plot(cbind(len_reads.df$length, cumsum(len_reads.df$reads)), main = "All data: Per ASV-length the number of reads",
     sub=basename(otu_tab_path), xlab = "ASV length", ylab="Cumulative sum reads")

hist(rowSums(otu_df), breaks = 100, main = "All data: Per OTU reads", sub=basename(otu_tab_path), 
     xlab = "Total number of reads")

hist(colSums(otu_df), breaks = 100, main = "All data: Per sample reads", sub=basename(otu_tab_path),
     xlab = "Total number of reads")

```


```{r obi-explore, eval=(TAX_TOOL=="OBITOOLS3"), fig.width=10, echo=F, fig.align='center'}
obitoolsID_check = tax_df_raw[row.names(otu_df), c(1), drop=F]

hist(obitoolsID_check$BEST_IDENTITY, breaks = 100,
     main = "All data: BEST_IDENTITY per OTU", sub=basename(tax_tab_path),
     xlab = "BEST_IDENTITY", ylab="number of OTUs")

ID_p = cbind(rowSums(otu_df),floor(obitoolsID_check$BEST_IDENTITY*100))
colnames(ID_p) = c("read_count", "BEST_ID_floor")
ID_p = aggregate(data=ID_p,  read_count ~ BEST_ID_floor, sum)


plot(ID_p$BEST_ID_floor, ID_p$read_count, main = "All data: BEST_IDENTITY per Reads",
     sub=basename(tax_tab_path) ,xlab = "BEST_IDENTITY", ylab="number of reads")
```


# Taxonomic annotation per SAMPLE/AMPLICON

```{r pss_plot-prep, echo=F}

tax_df_no_custom=tax_df[,!grepl("custom",colnames(tax_df))]
max_rank = function(x) if (sum(!is.na(x)) > 0) colnames(tax_df_no_custom)[max(which(!is.na(x)))] else "_root"
base_rank = apply(tax_df_no_custom, FUN=max_rank, MARGIN=1)
table(base_rank)

otu_df_rank = otu_df
otu_df_rank$max_tax_rank = base_rank[row.names(otu_df_rank)]

otu_df_rank = reshape2::melt(otu_df_rank, id.vars=c("max_tax_rank"), value.name = "read_count", variable.name="USI")
otu_df_rank = aggregate(data=otu_df_rank, read_count ~ max_tax_rank + USI, sum)

# rank on non-lowest level (aka how many reads not species level annot)
llv = tail(colnames(tax_df),1)
usilvl = otu_df_rank
if (length(unique(otu_df_rank$max_tax_rank)) != llv) {usilvl = aggregate(data=otu_df_rank[otu_df_rank$max_tax_rank != llv,], read_count ~ USI, sum)}
usilvl = usilvl$USI[order(usilvl$read_count, decreasing = F)]

otu_df_rank$USI = factor(otu_df_rank$USI, levels=usilvl)
otu_df_rank$max_tax_rank = factor(otu_df_rank$max_tax_rank, levels = rev(c("_root", colnames(tax_df_no_custom))))

pss_plot = ggplot(data=otu_df_rank, aes(x=USI, y=read_count, fill = max_tax_rank)) + geom_col(position = "stack") + 
  coord_flip() + ggtitle("Total sample reads by maximum taxonomic level", subtitle = paste0(basename(otu_tab_path), "\n",basename(tax_tab_path))) +
    theme(legend.text = element_text(size=12),
        legend.key.size=unit(0.4, "cm"),
        legend.position="top") +
  scale_y_continuous(expand = c(0,0))

sfigh = min(150, ncol(otu_df)*0.2)
```

```{r pss_plot, fig.height=sfigh, fig.width=20, echo=F, fig.align='center'}
pss_plot
```

- This might be quite small for the number of samples/amplicons
- Check the generated supplement `.pdf`

```{r pss_plot-ggsave, echo=F}
ggfname=paste0("operational-data-QC-report-supplement-fig.pdf")
ggoutp = file.path(params$OUT_DIR,ggfname)
print(ggfname)
ggsave(ggoutp, pss_plot, width = 25, height = ncol(otu_df)*0.2, limitsize = FALSE)
```


```{r obi3-ID-check, eval=(TAX_TOOL == "OBITOOLS3"), results='asis' ,echo=F}
ID_cutoff=1

cat(paste0("\n## [OBITOOLS] BEST_IDENTITY filter per sample: ID", ID_cutoff, ""))

obitoolsID_check$KEEP = FALSE
obitoolsID_check$KEEP[obitoolsID_check$BEST_IDENTITY >= ID_cutoff] = TRUE

obitoolsID_pl = reshape2::melt(cbind(otu_df, obitoolsID_check[, c("KEEP"), drop=F]), id.vars=c("KEEP"), value.name = "read_count", variable.name="USI")
obitoolsID_pl = aggregate(data=obitoolsID_pl, read_count ~ KEEP + USI, sum)
usi_rc_order = obitoolsID_pl[!obitoolsID_pl$KEEP,]
usi_rc_order = usi_rc_order$USI[order(usi_rc_order$read_count, decreasing = F)]
obitoolsID_pl$USI = factor(obitoolsID_pl$USI, levels=usi_rc_order)
percent_unknown = (sum(obitoolsID_pl$read_count[!obitoolsID_pl$KEEP]) / sum(obitoolsID_pl$read_count)) *100
obitoolsID_pl$KEEP = factor(obitoolsID_pl$KEEP, levels = rev(c(FALSE, TRUE)))

#make a barplot of the read counts per sample, this plot is "stacked" and will show the total number of reads, with the "unknown" fraction indicated in another color.
obi_ID_plot = ggplot(data=obitoolsID_pl, aes(x=USI, y=read_count, fill=KEEP)) +
  coord_flip() +
  geom_col(position="stack") +
  ggtitle(paste0("Number of reads with BEST_ID >=", ID_cutoff), subtitle = paste0(basename(otu_tab_path), "\n",basename(tax_tab_path), "\n", paste0("Percent unknown: ", round(percent_unknown,2), "%" )))   + 
  theme(legend.text = element_text(size=12),
        legend.key.size=unit(0.4, "cm"),
        legend.position="top") +
  scale_y_continuous(expand = c(0,0)) + scale_fill_brewer(palette="Dark2")

```

```{r obi3-ID-plot, eval=(TAX_TOOL == "OBITOOLS3"), fig.height=sfigh, fig.width=20, fig.align='center'}
print(obi_ID_plot)

ggfname1 =paste0("operational-data-QC-report-supplement-fig-bis.pdf")
ggoutp1 = file.path(params$OUT_DIR,ggfname1)
print(ggfname1)
ggsave(ggoutp1, obi_ID_plot, width = 25, height = ncol(otu_df)*0.2 , limitsize = FALSE)

```


```{r obi3-non1ID-check, eval=(TAX_TOOL == "OBITOOLS3"), echo=F, results='asis'}

cat(paste0("\n## [OBITOOLS] Inspect Abundant/Prevalent non ID",ID_cutoff, " OTUs"))

showN = 25

obitoolsID_check_rc = cbind(data.frame("total_reads" = rowSums(otu_df)), obitoolsID_check)
obitoolsID_check_rc$prevalence = apply(otu_df, function(x) sum(x > 0), MARGIN=1)
obitoolsID_check_rc$percent_reads = round(obitoolsID_check_rc$total_reads / sum(rowSums(otu_df) * 100), 10)

# select non cutoff OTUs
obitoolsID_check_rc = obitoolsID_check_rc[!obitoolsID_check_rc$KEEP, c(1,5,4,2)]

# add sequences and prelim taxon
obitoolsID_check_rc$taxonomy = tax_df_raw[row.names(obitoolsID_check_rc), c("SCIENTIFIC_NAME")]
obitoolsID_check_rc$nuc_seq = as.character(seqs[row.names(obitoolsID_check_rc)])


# print(tapply(obitoolsID_check_rc$total_reads, obitoolsID_check_rc$taxonomy, sum))

# Tables

#high read
max_rc_unk = obitoolsID_check_rc[order(obitoolsID_check_rc$total_reads, decreasing = T)[1:showN],]
kable(max_rc_unk, caption = paste0("non ID", ID_cutoff, " OTUs - ", showN, " Most Abundant"))

max_prev_unk = obitoolsID_check_rc[order(obitoolsID_check_rc$prevalence, decreasing = T)[1:showN],]
kable(max_prev_unk, caption = paste0("non ID", ID_cutoff, " OTUs - ", showN, " Most Prevalent"))

# plots

#//// after discussion
```



# Make Phyloseq object

- To work with Phyloseq, check the documentation:
  - https://joey711.github.io/phyloseq/

```{r make-ps}
# Make phyloseq object
ps = phyloseq(
  otu_table(otu_df, taxa_are_rows = TRUE),
  tax_table(as.matrix(tax_df)),
  sample_data(sample_sheet),
  seqs
)
ps 
```

## Check CONTROL samples/amplicons

```{r show-c, echo=F}
type.tb = table(sample_data(ps)$ANALYSE_TYPE)
type.tb

controls = names(type.tb)[names(type.tb) != "Metabarcoding"]

control_plots = as.list(controls, controls)
names(control_plots) = controls

csam.df = data.frame(sample_data(ps))
csam.df = csam.df[csam.df$ANALYSE_TYPE %in% controls,]

fH = max(length(unique(csam.df$TECHNICAL_REPLICATE_CODE)), 7)
fW = min(7,length(unique(csam.df$INDIVIDUAL_CODE))*0.7)
```

```{r plot-c, fig.width=fW, fig.height=fH, echo=F, fig.align='center'}
topx=20

for (i in 1:length(controls)){
  print(controls[i])
  maskcontrol = sample_data(ps)[["ANALYSE_TYPE"]] == controls[i]
  ps_control = prune_samples(maskcontrol, ps)
  
  if (sum(taxa_sums(ps_control)) < 1){
    cat(paste0(controls[i], " are all completely empty!"))
    next
  }
  
  ps_control = prune_taxa(rowSums(otu_table(ps_control)) > 0, ps_control)
  ps_control = tax_glom(ps_control, "species")
  ps_control
  
  # # create custom color
  txs = taxa_sums(ps_control)
  txs_keep = names(txs[order(txs, decreasing = T)][1:min(length(txs),topx)])
  txd = as.data.frame(tax_table(ps_control))
  txd$top_species = NA
  txd[txs_keep, "top_species"] = txd[txs_keep, "species"]

  tax_table(ps_control) = tax_table(as.matrix(txd))
  
  p_overview_control = plot_bar(ps_control, fill = "custom_taxon", x = "INDIVIDUAL_CODE") + 
  facet_grid(rows = vars(TECHNICAL_REPLICATE))+
  ggtitle(paste(controls[i], "- overview"), subtitle = PROJ_NAME)  +
    theme(legend.position="top") +
  scale_y_continuous(expand = c(0,0))
  print(p_overview_control)
  
  p_overview_control_species = plot_bar(ps_control, fill = "top_species", x = "INDIVIDUAL_CODE") + 
  facet_grid(rows = vars(TECHNICAL_REPLICATE))+
  ggtitle(paste(controls[i], "- top",topx,"taxa"), subtitle = PROJ_NAME) +
    theme(legend.position="top") +
  scale_y_continuous(expand = c(0,0)) 
  print(p_overview_control_species)
}
```


## Overview

```{r MB-samples-overview, echo=F}
### Inspect Samples ###
ps_samples = subset_samples(ps, ANALYSE_TYPE_CODE=="999")

# Look at top x ASVs (reduce for taxglom)
x_cutoff= 5000
topX = names(taxa_sums(ps_samples)[order(taxa_sums(ps_samples), decreasing = T)][1:min(ntaxa(ps_samples) ,x_cutoff)])

ps_topX = prune_taxa(topX, ps_samples)

ps_topX = tax_glom(ps_topX, "custom_taxon")
```

- Remove control samples
- Only look at top `r x_cutoff` top abundant ASVs (reads) to reduce data -> this amounts to `r round(sum(taxa_sums(ps_topX))/sum(taxa_sums(ps_samples)) *100, 2)`% of data
- group on custom taxonomy

```{r plot_bar, fig.height=25, fig.width=30, warning=F, echo=F, , fig.align='center'}
normalized_ps = transform_sample_counts(ps_topX, function(OTU) OTU/sum(OTU) * 100)
  
plot_bar(normalized_ps, fill = "custom_taxon", x="INDIVIDUAL_CODE") + 
  # coord_flip() +
  facet_grid(rows = vars(TECHNICAL_REPLICATE)) + 
  ggtitle(paste(PROJ_NAME)) + ylab('Relative abundance (%)') +  theme(legend.position = "bottom")

```


# Taxonomy based clustering metrics

We can use taxonomy information to (more or less) assess the clustering results.

- Ideally 1 cluster == 1 species -> check for number of OTUs per species annotation

Note that taxonomy is often not real "Ground-Truth":

- **Reference databases** have mistakes
- Taxonomic annotations **tool** can make false positive/negative taxonomic annotations
- **Noisy data** will lead to increased false positive annotations, picking up noise as signal. (=> data not enough filtered)
- Target **amplicon** might not resolve into species, or have intra-species variation.


```{r big-loop, echo=F}

# make plot input data
ps_select = prune_taxa(topX, ps_samples)
pstab = as.data.frame(tax_table(ps_select))
tax_levels=colnames(pstab)
pstab$total_count = taxa_sums(otu_table(ps_select))

# Parameters
topN_search = 15

# Containers
taxdata = data.frame()
pl_list = list()
pl_reads_list = list()
pl_otus_list = list()
pl_scatter_list = list()
tb_list= list()

# Make all plots/tables per taxonomic level
for (tax_rank in tax_levels){
  # Skip custom-taxon if present
  if (grepl("custom", tax_rank)){next}
  
  # Check number of OTU per taxrank
  tx_tab_i_total = data.frame(table(pstab[[tax_rank]], useNA = "ifany"))
  colnames(tx_tab_i_total) = c("Taxonomy", "nr_otus")
  tx_tab_i_total$Taxonomy = as.character(tx_tab_i_total$Taxonomy)
  tx_tab_i_total$Taxonomy[is.na(tx_tab_i_total$Taxonomy)] = "UNCLASSIFIED"
  
  # Check reads per taxrank
  temp_i = pstab[[tax_rank]]
  temp_i[is.na(temp_i)] = "UNCLASSIFIED"
  reads_df_i = aggregate(data=pstab, pstab$total_count ~ temp_i, sum)
  colnames(reads_df_i) = c("Taxonomy", "nr_reads_total")
  # Combine
  tx_tab_i_total = merge(tx_tab_i_total, reads_df_i)
  
  # select topN - most reads
  tx_tab_i_topN_reads = tx_tab_i_total[order(tx_tab_i_total$nr_reads_total, decreasing = T)[1:min(nrow(tx_tab_i_total),topN_search)],]
  tx_tab_i_topN_otus = tx_tab_i_total[order(tx_tab_i_total$nr_otus, decreasing = T)[1:min(nrow(tx_tab_i_total),topN_search)],]
  topN=topN_search
  tx_tab_i_topN = rbind(tx_tab_i_topN_reads, tx_tab_i_topN_otus)
  tx_tab_i_topN = tx_tab_i_topN[!duplicated(tx_tab_i_topN$Taxonomy),]

  rest_row = data.frame("Taxonomy" = paste0("OTHER_", toupper(tax_rank),"_sum"),
              "nr_otus"=sum(tx_tab_i_total$nr_otus[!row.names(tx_tab_i_total) %in% row.names(tx_tab_i_topN)]),
              "nr_reads_total"=sum(tx_tab_i_total$nr_reads_total[!row.names(tx_tab_i_total) %in% row.names(tx_tab_i_topN)]))
  
  if (rest_row$nr_reads_total > 0){
      tx_tab_i_topN = rbind(tx_tab_i_topN, rest_row)
  }

  tx_tab_i_topN$Taxonomy = as.character(tx_tab_i_topN$Taxonomy)

  tx_tab_i_topN = tx_tab_i_topN[tx_tab_i_topN$nr_reads_total > 0,]
  tx_tab_i_topN$nr_otus_rela = round(tx_tab_i_topN$nr_otus/sum(tx_tab_i_topN$nr_otus)*100, 2)
  tx_tab_i_topN$nr_reads_rela = round(tx_tab_i_topN$nr_reads_total/sum(tx_tab_i_topN$nr_reads_total)*100, 2)
  tx_tab_i_topN = tx_tab_i_topN[order(tx_tab_i_topN$nr_reads_total, decreasing = F),]
  
  tx_tab_i_topN$Taxonomy = factor(tx_tab_i_topN$Taxonomy, levels=tx_tab_i_topN$Taxonomy)

  myc = paste(nrow(tx_tab_i_topN),"taxa with most OTU|reads")

  # 1. Table
  kbt = kable(tx_tab_i_topN[order(tx_tab_i_topN$nr_otus, tx_tab_i_topN$nr_reads_total, decreasing = T),], caption = myc, col.names = c(tax_rank, "nr_OTU", "nr_total_reads", "OTU(%)", "reads(%)"),row.names = FALSE)
  tb_list[[tax_rank]] = kbt
  
  # 2. OTUs per taxon 
  default_cols <- scales::hue_pal()(length(levels(tx_tab_i_topN$Taxonomy)))
  default_cols[grepl("^UNCLASS", levels(tx_tab_i_topN$Taxonomy))] = "black"
  default_cols[grepl("^OTHER_", levels(tx_tab_i_topN$Taxonomy))] = "lightgrey"
  
  # not interesting for species
  pl1data=tx_tab_i_topN
  pl1cols=default_cols
  pl1myc = myc
  tax_order = tx_tab_i_topN$Taxonomy[order(tx_tab_i_topN$nr_otus, tx_tab_i_topN$nr_reads_total, decreasing = T)]
  if (grepl("^s", tax_rank)){
      pl1data = tx_tab_i_topN[!grepl("^OTHER|^UNCLASSIFIED", tx_tab_i_topN$Taxonomy),]
      pl1cols=scales::hue_pal()(length(levels(tx_tab_i_topN$Taxonomy)))
      # pl1myc = paste0(myc, "\nNOTE: only top ", topN, " shown")
  }
  
  pl1 = ggplot(data=pl1data, aes(x = Taxonomy, y=nr_otus, , fill=Taxonomy)) + 
  geom_col() + ggtitle(paste(tax_rank, " - Number of OTUs"), subtitle = paste0("Shown: ", pl1myc)) + 
  coord_flip() + scale_fill_manual(values = pl1cols) +  theme_minimal()
  
  pl_otus_list[[tax_rank]] = pl1
  
  # 2. Read counts per taxon
  pl2 = ggplot(data=tx_tab_i_topN, aes(x = Taxonomy, y=nr_reads_total, fill=Taxonomy)) + 
  geom_col() + ggtitle(paste(tax_rank, " - Total number of reads"), subtitle = paste0("Shown: ", myc)) + 
  coord_flip() + scale_fill_manual(values = default_cols) + theme_minimal() 
  # Store in list
  pl_reads_list[[tax_rank]] = pl2
  
  # 3. Scatter readcounts ~ nr_otus
  sc_pl = tx_tab_i_topN[!grepl("^OTHER|^UNCLASSIFIED", tx_tab_i_topN$Taxonomy),]
  pl3 = ggplot(data=sc_pl[!is.na(sc_pl$Taxonomy),], aes(x = nr_reads_total , y= nr_otus, color=Taxonomy, label=Taxonomy)) + 
  ggtitle(paste(tax_rank, " - within tax-rank OTU spread"), subtitle = paste0("Shown: ", myc)) + coord_flip() + geom_label(size=2, alpha=0.5) +
    theme(legend.position="bottom")
  pl_scatter_list[[tax_rank]] = pl3
  
  # 4. unique OTUs per taxon
  # This is only informative for species
  if (grepl("^s", tax_rank)){
    otu_count_tab_i = data.frame(table(table(pstab[[tax_rank]])))
    otu_count_tab_i$Var1 = as.numeric(as.character(otu_count_tab_i$Var1))
    colnames(otu_count_tab_i) = c("nr_otus", "nr_taxa")
    otu_count_tab_i$type = "Annotated"
    
    otu_count_tab_i$nr_otus[otu_count_tab_i$nr_otus > 100] = 101
    pl =   ggplot(data=otu_count_tab_i, aes(x = nr_otus, y=nr_taxa)) + geom_col() + 
    ggtitle(paste0(tax_rank,  " - #OTU per taxon"), subtitle = "Note that unclassified OTUs are NOT included") + xlab("#OTUs per taxon (100+ is grouped)") +
    ylab("#taxa")
    
    pl_list[[tax_rank]] = pl
  }
  
}

```


## Taxonomy in function of OTU variation

- OTU variation = number of distinct OTUs per taxon
- For **non-species taxonomic ranks** this _might_ indicate the diversity of the taxon in the dataset
- For the **species level** this indicates intra-specific variation in the amplicon (either artifacts/errors/noise or biological)
  - When a lot of amplicons exist for a single species this might require more detailed investigation.

```{r pl_otus_list, fig.width=10, fig.height=5, fig.align='center', echo=F}
pl_otus_list
```

### Number of OTUs with unique species annotation

- The more OTUs have a unique species annotation the "better" the amplicon performed for your data.

```{r pl_list, echo=F, fig.align='center'}
pl_list
```


## Taxonomy in function of TOTAL number of reads

- Highlight the taxa most supported by number of reads.

```{r pl_reads_list, fig.width=10, fig.height=5, echo=F, fig.align='center'}

pl_reads_list

```

## Taxonomy in function of OTU variation and TOTAL number of reads

- Combine the OTU variation with the number of reads for each taxon in 2D

```{r pl_scatter_list, fig.width=10, fig.height=8, echo=F, fig.align='center'}
pl_scatter_list$species
```


## Taxonomy tables

- Tables with most occurring taxonomies on each rank
- Most occurring is defined in OTU variation and in number of reads
- Note that not all taxons with a lot of reads are also diverse, and vice versa
  - Taxonomic resolution of amplicon can be different ~ taxonomic group

```{r tb_list, echo=F, results='asis'}

for (ttab in tb_list){
  print(ttab)
}

```

# More questions?

You can quite easily load the data into an R session using gendiver:

- https://github.com/inbo/gendiver

Use the following links/paths:

NOTE: you can replace PATHS with the URL of files on the GDRIVE!

- `params$SAMPLE_SHEET_URL`: `r params$SAMPLE_SHEET_URL`
- `params$OTU_PATH`: `r params$OTU_PATH`
- `params$TAX_PATH`: `r params$TAX_PATH`
- `params$FASTA_PATH`: `r params$FASTA_PATH`

Examples to load in data (see also INPUT segment of this report):

```{r example-load, eval=F, echo = T}
# Load custom R-pkg
library(gendiver)

my_sample_data = read.sample_sheet_lab(params$SAMPLE_SHEET_URL)

my_otu_tab = read.otu_table_vsearch(params$OTU_PATH)

my_tax_tab = read.taxonomy_sintax(params$TAX_PATH)

my_seqs = read.fastx_gdrive(params$FASTA_PATH)

```

