---
title: "Operational Data Report"
author: 'Nick Dillen'
date: "`r Sys.Date()`"
output:
  # pdf_document: 
  #   toc: true
  #   fig_caption: true
  html_document:
    code_folding: hide
    highlight: haddock
    number_sections: false
    fig_caption: true
    theme: sandstone
    toc: true
    toc_float: true
urlcolor: blue
params:
  SAMPLE_SHEET_URL: "URL-sample-sheet"
  TAX_PATH: "TAX-result"
  OTU_PATH: "OTU-table"
  FASTA_PATH: "FASTA-path"
  OUT_DIR: "."
---

```{r setup, echo=FALSE,}
options(width = 80)
```


# Operational Data report

- Investigate basic statistic of combined **Clustering** and **Taxonomy** for a certain dataset
- Interpret and evaluate **non-experiment** samples/amplicons

## Input data

- Input lab sample sheet: `r params$SAMPLE_SHEET_URL`
- Input OTU table: `r params$OTU_PATH`
- Input Taxonomy file: `r params$TAX_PATH`
- Input FASTA: `r params$FASTA_PATH`

```{r read-param, echo=FALSE, eval=T, message=F}
library(gendiver)
library(knitr)
library(ggplot2)
library(phyloseq)
library(Biostrings)

googlesheets4::gs4_auth(T)
googledrive::drive_auth(T)

otu_tab_path = params$OTU_PATH
tax_tab_path = params$TAX_PATH
fasta_path = params$FASTA_PATH

# Check file location
if (! file.exists(otu_tab_path)) {
  otu_tab_path = googledrive::drive_get(otu_tab_path)
  otu_tab_path = otu_tab_path$name
}

if (! file.exists(tax_tab_path)) {
  tax_tab_path = googledrive::drive_get(tax_tab_path)
  tax_tab_path = tax_tab_path$name
}

if (! file.exists(fasta_path)) {
  fasta_path = googledrive::drive_get(fasta_path)
  fasta_path = fasta_path$name
}

PROJ_NAME = gsub("\\..*", "", basename(otu_tab_path))
SS_NAME = googledrive::drive_get(params$SAMPLE_SHEET_URL)$name


SUPP_DIR = file.path(params$OUT_DIR, "supplementary")
dir.create(SUPP_DIR)
```

- Used lab sample sheet: `r SS_NAME`
- Used OTU table: `r basename(otu_tab_path)`
- Used Taxonomy file: `r basename(tax_tab_path)`
- Used sequence file: `r basename(fasta_path)`


\break

# Read data

```{r read.opdata, echo=F, warning=F, message=T}

sample_sheet = read.sample_sheet_lab(params$SAMPLE_SHEET_URL)

seqs = read.fastx_gdrive(params$FASTA_PATH)
names(seqs) = gsub(";size=\\d*", "", names(seqs))

if (grepl("obitools", otu_tab_path, ignore.case = T)){
  otu_df <- read.otu_table_obitools3(params$OTU_PATH)
} else {
  #Read OTU table and tax table, and explore data
  otu_df <- read.otu_table_vsearch(params$OTU_PATH)
}

if (grepl("sintax", tax_tab_path, ignore.case = T)){
  tax_df =  read.taxonomy_sintax(params$TAX_PATH, clean_tax = T, add_custom = T)
  TAX_TOOL="SINTAX"
} else if(grepl("idtaxa", tax_tab_path, ignore.case = T)){
  #Read OTU table and tax table, and explore data
  tax_df <- read.taxonomy_idtaxa(params$TAX_PATH)
  TAX_TOOL="IDTAXA"
} else if(grepl("obitools3", tax_tab_path, ignore.case = T)){
  #Read OTU table and tax table, and explore data
  tax_df_raw <- read.taxonomy_obitools3(params$TAX_PATH)
  tax_df = select.taxonomy_obitools3(tax_df_raw, min_BEST_ID = 0, fillNA = F)
  
  # seqs = seqs[row.names(tax_df)]
  TAX_TOOL="OBITOOLS3"

} 
  
```

```{r raw-ps}
ps = phyloseq(
  otu_table(otu_df, taxa_are_rows = TRUE),
  tax_table(as.matrix(tax_df)),
  sample_data(sample_sheet),
  seqs
)
ps
```

\break

# Export datasets

```{r export-edna-water, eval=(TAX_TOOL == "OBITOOLS3")}
datadir = file.path(params$OUT_DIR, "data_export")
dir.create(datadir)
gendiver::export.data_sets(
  sample_sheet_df = sample_sheet, otu_df = otu_df, tax_df = tax_df_raw, out_path = datadir)
```

```{r export-ps}
saveRDS(ps, file.path(paste0(PROJ_NAME, "_phyloseq_raw.rds" )))
```

# Basic exploration of Operational Data

```{r basic-explore, fig.width=10, echo=F, fig.align='center'}

gendiver::qcplot.basic_exploration(ps)

```


```{r obi-explore, eval=(TAX_TOOL=="OBITOOLS3"), fig.width=10, echo=F, fig.align='center'}
obitoolsID_check = tax_df_raw[row.names(otu_df), c(1), drop=F]

hist(obitoolsID_check$BEST_IDENTITY, breaks = 100,
     main = "All data: BEST_IDENTITY per OTU", sub=basename(tax_tab_path),
     xlab = "BEST_IDENTITY", ylab="number of OTUs")

ID_p = cbind(rowSums(otu_df),floor(obitoolsID_check$BEST_IDENTITY*100))
colnames(ID_p) = c("read_count", "BEST_ID_floor")
ID_p = aggregate(data=ID_p,  read_count ~ BEST_ID_floor, sum)


plot(ID_p$BEST_ID_floor, ID_p$read_count, main = "All data: BEST_IDENTITY per Reads",
     sub=basename(tax_tab_path) ,xlab = "BEST_IDENTITY", ylab="number of reads")
```

```{r obi3-ID-check, eval=(TAX_TOOL == "OBITOOLS3"), results='asis' ,echo=F}
ID_cutoff=1

cat(paste0("\n## [OBITOOLS] BEST_IDENTITY filter per sample: ID", ID_cutoff, ""))

obitoolsID_check$KEEP = FALSE
obitoolsID_check$KEEP[obitoolsID_check$BEST_IDENTITY >= ID_cutoff] = TRUE

obitoolsID_pl = reshape2::melt(cbind(otu_df, obitoolsID_check[, c("KEEP"), drop=F]), id.vars=c("KEEP"), value.name = "read_count", variable.name="USI")
obitoolsID_pl = aggregate(data=obitoolsID_pl, read_count ~ KEEP + USI, sum)
usi_rc_order = obitoolsID_pl[!obitoolsID_pl$KEEP,]
usi_rc_order = usi_rc_order$USI[order(usi_rc_order$read_count, decreasing = F)]
obitoolsID_pl$USI = factor(obitoolsID_pl$USI, levels=usi_rc_order)
percent_unknown = (sum(obitoolsID_pl$read_count[!obitoolsID_pl$KEEP]) / sum(obitoolsID_pl$read_count)) *100
obitoolsID_pl$KEEP = factor(obitoolsID_pl$KEEP, levels = rev(c(FALSE, TRUE)))

#make a barplot of the read counts per sample, this plot is "stacked" and will show the total number of reads, with the "unknown" fraction indicated in another color.
obi_ID_plot = ggplot(data=obitoolsID_pl, aes(x=USI, y=read_count, fill=KEEP)) +
  coord_flip() +
  geom_col(position="stack") +
  ggtitle(paste0("Number of reads with BEST_ID >=", ID_cutoff), subtitle = paste0(basename(otu_tab_path), "\n",basename(tax_tab_path), "\n", paste0("Percent unknown: ", round(percent_unknown,2), "%" )))   + 
  theme(legend.text = element_text(size=12),
        legend.key.size=unit(0.4, "cm"),
        legend.position="top") +
  scale_y_continuous(expand = c(0,0)) + scale_fill_brewer(palette="Dark2")

sfigh = min(150, ncol(otu_df)*0.2)
```

```{r obi3-ID-plot, eval=(TAX_TOOL == "OBITOOLS3"), fig.height=sfigh, fig.width=20, fig.align='center'}
print(obi_ID_plot)

ggfname1 =paste0("obitools-ID-coverage.pdf")
ggoutp1 = file.path(SUPP_DIR, ggfname1)
print(ggfname1)
ggsave(ggoutp1, obi_ID_plot, width = 25, height = ncol(otu_df)*0.2 , limitsize = FALSE)
invisible(gc())
```


```{r obi3-non1ID-check, eval=(TAX_TOOL == "OBITOOLS3"), echo=F, results='asis'}

cat(paste0("\n## [OBITOOLS] Inspect Abundant/Prevalent non ID",ID_cutoff, " OTUs"))

showN = 25

obitoolsID_check_rc = cbind(data.frame("total_reads" = rowSums(otu_df)), obitoolsID_check)
obitoolsID_check_rc$prevalence = apply(otu_df, function(x) sum(x > 0), MARGIN=1)
obitoolsID_check_rc$percent_reads = round(obitoolsID_check_rc$total_reads / sum(rowSums(otu_df) * 100), 10)

# select non cutoff OTUs
obitoolsID_check_rc = obitoolsID_check_rc[!obitoolsID_check_rc$KEEP, c(1,5,4,2)]

# add sequences and prelim taxon
obitoolsID_check_rc$taxonomy = tax_df_raw[row.names(obitoolsID_check_rc), c("SCIENTIFIC_NAME")]
obitoolsID_check_rc$nuc_seq = as.character(seqs[row.names(obitoolsID_check_rc)])


# print(tapply(obitoolsID_check_rc$total_reads, obitoolsID_check_rc$taxonomy, sum))

# Tables

#high read
max_rc_unk = obitoolsID_check_rc[order(obitoolsID_check_rc$total_reads, decreasing = T)[1:showN],]
kable(max_rc_unk, caption = paste0("non ID", ID_cutoff, " OTUs - ", showN, " Most Abundant"))

max_prev_unk = obitoolsID_check_rc[order(obitoolsID_check_rc$prevalence, decreasing = T)[1:showN],]
kable(max_prev_unk, caption = paste0("non ID", ID_cutoff, " OTUs - ", showN, " Most Prevalent"))

# plots

#//// after discussion
```

```{r obi3-tax-df, eval=(TAX_TOOL == "OBITOOLS3")}
# remake tax_df as all ID1 ASVs
tax_df = select.taxonomy_obitools3(tax_df_raw, min_BEST_ID = ID_cutoff, fillNA = F)

rm_unk_reads = sum(colSums(otu_df[row.names(tax_df),]))
rm_unk = 100*(1 - rm_unk_reads/ sum(colSums(otu_df)))

message(paste0("[INFO] This removes ", rm_unk_reads, " (",round(rm_unk, 2), "%) reads"))

```


# Make Phyloseq object

- To work with Phyloseq, check the documentation:
  - https://joey711.github.io/phyloseq/

```{r make-ps}
# Make phyloseq object
ps = phyloseq(
  otu_table(otu_df, taxa_are_rows = TRUE),
  tax_table(as.matrix(tax_df)),
  sample_data(sample_sheet),
  seqs
)
ps 

rm(otu_df, tax_df, seqs)
invisible(gc())
```

# Taxonomic annotation per SAMPLE/AMPLICON

```{r pss_plot-prep, echo=F}

pss_plot = qcplot.sample_barplot_taxonomic_coverage(ps)

invisible(gc())

```

```{r pss_plot, fig.height=sfigh, fig.width=20, echo=F, fig.align='center'}
pss_plot
```

- This might be quite small for the number of samples/amplicons
- Check the generated supplement `.pdf`

```{r pss_plot-ggsave, echo=F}
ggfname=paste0("taxonomic-annotation-coverage.pdf")
ggoutp = file.path(SUPP_DIR,ggfname)
print(ggfname)
ggsave(ggoutp, pss_plot, width = 25, height = sfigh, limitsize = FALSE)
rm(pss_plot)
invisible(gc())
```


# CONTROL samples/amplicons

```{r show-c, echo=F}
type.tb = table(sample_data(ps)$ANALYSE_TYPE)
if (length(type.tb) < 1){type.tb = table(sample_data(ps)$SAMPLE_TYPE)}
type.tb

controls = names(type.tb)[names(type.tb) != "Metabarcoding" & names(type.tb) != "eDNA"]

control_plots = as.list(controls, controls)
names(control_plots) = controls

csam.df = data.frame(sample_data(ps))
csam.df = csam.df[csam.df$ANALYSE_TYPE %in% controls,]

fH = max(length(unique(csam.df$TECHNICAL_REPLICATE_CODE))*2, 8)
fW = max(15 , min(25,length(unique(csam.df$INDIVIDUAL_CODE))*0.7))
```

```{r plot-c, fig.width=fW, fig.height=fH, echo=F, fig.align='center'}

if (length(control_plots) > 0){
  
  topx=20

  for (i in 1:length(controls)){
    print(controls[i])
    maskcontrol = sample_data(ps)[["ANALYSE_TYPE"]] == controls[i]
    ps_control = prune_samples(maskcontrol, ps)
    
    if (sum(taxa_sums(ps_control)) < 1){
      cat(paste0(controls[i], " are all completely empty!"))
      next
    }
    
    ps_control = prune_taxa(rowSums(otu_table(ps_control)) > 0, ps_control)
    
    # find highst tax with no NA
    tax_na = colSums(!is.na(tax_table(ps_control)))
    tax_level = names(tax_na)[max(which(tax_na > 0))]
    
    ps_control = tax_glom(ps_control, tax_level)
    ps_control
    
    # # create custom color
    txs = taxa_sums(ps_control)
    txs_keep = names(txs[order(txs, decreasing = T)][1:min(length(txs),topx)])
    txd = as.data.frame(tax_table(ps_control))
    txd$top_taxa = NA
    txd[txs_keep, "top_taxa"] = txd[txs_keep, tax_level]
  
    tax_table(ps_control) = tax_table(as.matrix(txd))
    
    p_overview_control = plot_bar(ps_control, fill = "custom_taxon", x = "INDIVIDUAL_CODE") + 
    facet_grid(rows = vars(TECHNICAL_REPLICATE))+
    ggtitle(paste(controls[i], "- overview"), subtitle = PROJ_NAME)  +
      theme(legend.position="top") +
    scale_y_continuous(expand = c(0,0))
    print(p_overview_control)
    
    p_overview_control_species = plot_bar(ps_control, fill = "top_taxa", x = "INDIVIDUAL_CODE") + 
    facet_grid(rows = vars(TECHNICAL_REPLICATE))+
    ggtitle(paste(controls[i], "- top",topx,"taxa"), subtitle = PROJ_NAME) +
      theme(legend.position="top", legend.text = element_text(size=8)) +
    scale_y_continuous(expand = c(0,0)) 
    print(p_overview_control_species)
    }
  
} else {
  print("No CONTROLS detected !")
}


```


# Overview of PS object

- redo basic analysis (for obitools data this is now a pruned set)

```{r qc_vis_basic, echo=F, eval=F}
### QC visualisation of data ###
qcplot.basic_exploration(ps)
```

## Plate column analysis

- To check for column swaps

```{r qc_vis_columns, echo=F, warning=F, eval=(TAX_TOOL == "OBITOOLS3")}
qcplot.plate_column_asv_analysis(ps)
```

## Plate readcount heatmap

```{r}

summary(sample_sums(ps))

omit_cutoff = round(median(sample_sums(ps)) * 0.01)
```


- I choose the cutoff as 1% of median sample (`r omit_cutoff`)


### All data

```{r, fig.width=15, fig.height=15}


qcplot.plate_heatmap_readcount(ps, omit_cutoff = omit_cutoff)

qcplot.plate_heatmap_readcount(ps, omit_cutoff = omit_cutoff, transform = "log")

```

### Inspect the low read "failed" samples

- samples with less than `r omit_cutoff` (1% of median) reads

```{r failed-samples-create}
failedsamples = prune_samples(sample_sums(ps) < omit_cutoff, ps)
failedsamples = prune_taxa(taxa_sums(failedsamples) > 0, failedsamples)
```


```{r failed-samples-view, fig.width=15, fig.height=8}
make.taxa_barplot(failedsamples, taxa_rank = "species", taxa_excl = NA) + ggtitle(paste0('Samples with < ', omit_cutoff, ' (1% of median) reads'))
```


Check this for the custom taxons as well. 

How do these taxa spread over the plates, is this as expected?

I use 1% of the previously determined "minimum reads for a successful sample" as the minimum reads of the selected taxa for a sample to be plotted (`r round(omit_cutoff * 0.1) `).


```{r zoom-heatmap}
zh_list = list()

ttdf = data.frame(tax_table(ps)[, "custom_taxon"])

ct_list = unique(ttdf$custom_taxon)

for (cct in ct_list){
  ps_SELECT = prune_taxa(ttdf$custom_taxon == cct, ps)
  taxa_omit = round(omit_cutoff * 0.1)
  zpl = qcplot.plate_heatmap_readcount(ps_SELECT, omit_cutoff = taxa_omit) 
  
  if (!is.null(zpl)){
    zpl = zpl + ggtitle(paste0("Read counts per well, only ", ID_cutoff, "ID reads, taxon=", cct ))

  } else {
    zpl=paste("No samples with more than", taxa_omit, cct, "reads" )
  }
  
  zh_list[[cct]] = zpl
  
}

```

```{r zh-plot, fig.width=15, fig.height=10}

zh_list

```



To get an overview of amplicons on the plate, and if the patterns match the expectations we can plot the most abundant ASV for each well


```{r fig.width=15, fig.height=15}

# summary(taxa_sums(ps))

qcplot.plate_heatmap_toptaxa(ps, omit_cutoff = 100)
```

# Experiment data

```{r sample_ps, echo=F}
### Inspect Samples ###
ps_samples = subset_samples(ps, ANALYSE_TYPE_CODE=="999")
```


```{r MB-samples-overview, echo=F}
# Look at top x ASVs (reduce for taxglom)
x_cutoff= 5000
topX = names(taxa_sums(ps_samples)[order(taxa_sums(ps_samples), decreasing = T)][1:min(ntaxa(ps_samples) ,x_cutoff)])

ps_topX = prune_taxa(topX, ps_samples)

ps_topX = tax_glom(ps_topX, "custom_taxon")
```

- Remove control samples
- Only look at top `r x_cutoff` top abundant ASVs (reads) to reduce data -> this amounts to `r round(sum(taxa_sums(ps_topX))/sum(taxa_sums(ps_samples)) *100, 2)`% of data
- group on custom taxonomy


```{r plot_bar, fig.height=25, fig.width=30, warning=F, echo=F, , fig.align='center'}
normalized_ps = transform_sample_counts(ps_topX, function(OTU) OTU/sum(OTU) * 100)
  
plot_bar(normalized_ps, fill = "custom_taxon", x="INDIVIDUAL_CODE") + 
  # coord_flip() +
  facet_grid(rows = vars(TECHNICAL_REPLICATE)) + 
  ggtitle(paste(PROJ_NAME)) + ylab('Relative abundance (%)') +  theme(legend.position = "bottom")
rm(normalized_ps)
invisible(gc())
```


# OTU abundance

REMOVE SEQUENCE VARIANTS WITH LOW OVERALL ABUNDANCE


Sequence variants that occur in very minor amounts are usually not interesting.

We check how much read data we loose when we remove the sequences with number of reads lower than a certain threshold by constructing a plot.

The plot shows how many reads are still left when using a certain threshold

```{r}

rownumber<-1
cutoff<-seq(10, 100000, by=10)    #we vary the cutoff from 10 to 100k in steps of 10
plotdata<-data.frame("cutoff" = cutoff,
                     "nr_reads" = NA,
                     "nr_otu" = NA)
otu_tot_data = taxa_sums(ps)

for (i in 1:nrow(plotdata)){
  plotdata[i, 2] = sum(otu_tot_data[otu_tot_data > plotdata$cutoff[i]])
  plotdata[i, 3] = sum(otu_tot_data > plotdata$cutoff[i])
}

plotdata$nr_reads_percent = 100 * plotdata$nr_reads / sum(otu_tot_data)
plotdata$nr_otu_percent = 100 * plotdata$nr_otu / length(otu_tot_data)

ggplot(data=plotdata) +
  geom_line(aes(x=cutoff,y=nr_reads_percent)) + ggtitle("Reads remaining per minR") +
  labs(x="Minimum Reads Per Taxa filtering cutoff (minR)",y="Percentage of reads kept")

ggplot(data=plotdata) +
  geom_line(aes(x=cutoff,y=nr_otu_percent)) + ggtitle("OTUs remaining per minR") +
  labs(x="Minimum Reads Per Taxa filtering cutoff (minR)",y="Percentage of OTUs kept")
```


- These plots can be an inspiration to choose the actual minR for data analysis

# Project data

- split the data in projects and visualize
- **These plots can be found in a supplementary folder alongside this report!** 
- (`r SUPP_DIR`)

```{r export-ppbar-plots, warning=F}

make.default_project_barplots(ps, RRA=F, out_path = file.path(SUPP_DIR, "project_barplots"))
make.default_project_barplots(ps, RRA=T, out_path = file.path(SUPP_DIR, "project_barplots_RRA"))
```


# Taxonomy based clustering metrics

We can use taxonomy information to (more or less) assess the clustering results.

- Ideally 1 cluster == 1 species -> check for number of OTUs per species annotation

Note that taxonomy is often not real "Ground-Truth":

- **Reference databases** have mistakes
- Taxonomic annotations **tool** can make false positive/negative taxonomic annotations
- **Noisy data** will lead to increased false positive annotations, picking up noise as signal. (=> data not enough filtered)
- Target **amplicon** might not resolve into species, or have intra-species variation.


```{r big-loop, echo=F}

# make plot input data
ps_select = prune_taxa(topX, ps_samples)
pstab = as.data.frame(tax_table(ps_select))
tax_levels=colnames(pstab)
pstab$total_count = taxa_sums(otu_table(ps_select))

# Parameters
topN_search = 15

# Containers
taxdata = data.frame()
pl_list = list()
pl_reads_list = list()
pl_otus_list = list()
pl_scatter_list = list()
tb_list= list()

# Make all plots/tables per taxonomic level
for (tax_rank in tax_levels){
  # Skip custom-taxon if present
  if (grepl("custom", tax_rank)){next}
  
  # Check number of OTU per taxrank
  tx_tab_i_total = data.frame(table(pstab[[tax_rank]], useNA = "ifany"))
  colnames(tx_tab_i_total) = c("Taxonomy", "nr_otus")
  tx_tab_i_total$Taxonomy = as.character(tx_tab_i_total$Taxonomy)
  tx_tab_i_total$Taxonomy[is.na(tx_tab_i_total$Taxonomy)] = "UNCLASSIFIED"
  
  # Check reads per taxrank
  temp_i = pstab[[tax_rank]]
  temp_i[is.na(temp_i)] = "UNCLASSIFIED"
  reads_df_i = aggregate(data=pstab, pstab$total_count ~ temp_i, sum)
  colnames(reads_df_i) = c("Taxonomy", "nr_reads_total")
  # Combine
  tx_tab_i_total = merge(tx_tab_i_total, reads_df_i)
  
  # select topN - most reads
  tx_tab_i_topN_reads = tx_tab_i_total[order(tx_tab_i_total$nr_reads_total, decreasing = T)[1:min(nrow(tx_tab_i_total),topN_search)],]
  tx_tab_i_topN_otus = tx_tab_i_total[order(tx_tab_i_total$nr_otus, decreasing = T)[1:min(nrow(tx_tab_i_total),topN_search)],]
  topN=topN_search
  tx_tab_i_topN = rbind(tx_tab_i_topN_reads, tx_tab_i_topN_otus)
  tx_tab_i_topN = tx_tab_i_topN[!duplicated(tx_tab_i_topN$Taxonomy),]

  rest_row = data.frame("Taxonomy" = paste0("OTHER_", toupper(tax_rank),"_sum"),
              "nr_otus"=sum(tx_tab_i_total$nr_otus[!row.names(tx_tab_i_total) %in% row.names(tx_tab_i_topN)]),
              "nr_reads_total"=sum(tx_tab_i_total$nr_reads_total[!row.names(tx_tab_i_total) %in% row.names(tx_tab_i_topN)]))
  
  if (rest_row$nr_reads_total > 0){
      tx_tab_i_topN = rbind(tx_tab_i_topN, rest_row)
  }

  tx_tab_i_topN$Taxonomy = as.character(tx_tab_i_topN$Taxonomy)

  tx_tab_i_topN = tx_tab_i_topN[tx_tab_i_topN$nr_reads_total > 0,]
  tx_tab_i_topN$nr_otus_rela = round(tx_tab_i_topN$nr_otus/sum(tx_tab_i_topN$nr_otus)*100, 2)
  tx_tab_i_topN$nr_reads_rela = round(tx_tab_i_topN$nr_reads_total/sum(tx_tab_i_topN$nr_reads_total)*100, 2)
  tx_tab_i_topN = tx_tab_i_topN[order(tx_tab_i_topN$nr_reads_total, decreasing = F),]
  
  tx_tab_i_topN$Taxonomy = factor(tx_tab_i_topN$Taxonomy, levels=tx_tab_i_topN$Taxonomy)

  myc = paste(nrow(tx_tab_i_topN),"taxa with most OTU|reads")

  # 1. Table
  kbt = kable(tx_tab_i_topN[order(tx_tab_i_topN$nr_otus, tx_tab_i_topN$nr_reads_total, decreasing = T),], caption = myc, col.names = c(tax_rank, "nr_OTU", "nr_total_reads", "OTU(%)", "reads(%)"),row.names = FALSE)
  tb_list[[tax_rank]] = kbt
  
  # 2. OTUs per taxon 
  default_cols <- scales::hue_pal()(length(levels(tx_tab_i_topN$Taxonomy)))
  default_cols[grepl("^UNCLASS", levels(tx_tab_i_topN$Taxonomy))] = "black"
  default_cols[grepl("^OTHER_", levels(tx_tab_i_topN$Taxonomy))] = "lightgrey"
  
  # not interesting for species
  pl1data=tx_tab_i_topN
  pl1cols=default_cols
  pl1myc = myc
  tax_order = tx_tab_i_topN$Taxonomy[order(tx_tab_i_topN$nr_otus, tx_tab_i_topN$nr_reads_total, decreasing = T)]
  if (grepl("^s", tax_rank)){
      pl1data = tx_tab_i_topN[!grepl("^OTHER|^UNCLASSIFIED", tx_tab_i_topN$Taxonomy),]
      pl1cols=scales::hue_pal()(length(levels(tx_tab_i_topN$Taxonomy)))
      # pl1myc = paste0(myc, "\nNOTE: only top ", topN, " shown")
  }
  
  pl1 = ggplot(data=pl1data, aes(x = Taxonomy, y=nr_otus, , fill=Taxonomy)) + 
  geom_col() + ggtitle(paste(tax_rank, " - Number of OTUs"), subtitle = paste0("Shown: ", pl1myc)) + 
  coord_flip() + scale_fill_manual(values = pl1cols) +  theme_minimal()
  
  pl_otus_list[[tax_rank]] = pl1
  
  # 2. Read counts per taxon
  pl2 = ggplot(data=tx_tab_i_topN, aes(x = Taxonomy, y=nr_reads_total, fill=Taxonomy)) + 
  geom_col() + ggtitle(paste(tax_rank, " - Total number of reads"), subtitle = paste0("Shown: ", myc)) + 
  coord_flip() + scale_fill_manual(values = default_cols) + theme_minimal() 
  # Store in list
  pl_reads_list[[tax_rank]] = pl2
  
  # 3. Scatter readcounts ~ nr_otus
  sc_pl = tx_tab_i_topN[!grepl("^OTHER|^UNCLASSIFIED", tx_tab_i_topN$Taxonomy),]
  pl3 = ggplot(data=sc_pl[!is.na(sc_pl$Taxonomy),], aes(x = nr_reads_total , y= nr_otus, color=Taxonomy, label=Taxonomy)) + 
  ggtitle(paste(tax_rank, " - within tax-rank OTU spread"), subtitle = paste0("Shown: ", myc)) + coord_flip() + geom_label(size=2, alpha=0.5) +
    theme(legend.position="bottom")
  pl_scatter_list[[tax_rank]] = pl3
  
  # 4. unique OTUs per taxon
  # This is only informative for species
  if (grepl("^s", tax_rank)){
    otu_count_tab_i = data.frame(table(table(pstab[[tax_rank]])))
    otu_count_tab_i$Var1 = as.numeric(as.character(otu_count_tab_i$Var1))
    colnames(otu_count_tab_i) = c("nr_otus", "nr_taxa")
    otu_count_tab_i$type = "Annotated"
    
    otu_count_tab_i$nr_otus[otu_count_tab_i$nr_otus > 100] = 101
    pl =   ggplot(data=otu_count_tab_i, aes(x = nr_otus, y=nr_taxa)) + geom_col() + 
    ggtitle(paste0(tax_rank,  " - #OTU per taxon"), subtitle = "Note that unclassified OTUs are NOT included") + xlab("#OTUs per taxon (100+ is grouped)") +
    ylab("#taxa")
    
    pl_list[[tax_rank]] = pl
  }
  
}

```


## Taxonomy in function of OTU variation

- OTU variation = number of distinct OTUs per taxon
- For **non-species taxonomic ranks** this _might_ indicate the diversity of the taxon in the dataset
- For the **species level** this indicates intra-specific variation in the amplicon (either artifacts/errors/noise or biological)
  - When a lot of amplicons exist for a single species this might require more detailed investigation.

```{r pl_otus_list, fig.width=10, fig.height=5, fig.align='center', echo=F}
pl_otus_list
rm(pl_otus_list)
invisible(gc())
```

### Number of OTUs with unique species annotation

- The more OTUs have a unique species annotation the "better" the amplicon performed for your data.

```{r pl_list, echo=F, fig.align='center'}
pl_list
rm(pl_list)
invisible(gc())
```


## Taxonomy in function of TOTAL number of reads

- Highlight the taxa most supported by number of reads.

```{r pl_reads_list, fig.width=10, fig.height=5, echo=F, fig.align='center'}

pl_reads_list
rm(pl_reads_list)
invisible(gc())
```

## Taxonomy in function of OTU variation and TOTAL number of reads

- Combine the OTU variation with the number of reads for each taxon in 2D

```{r pl_scatter_list, fig.width=10, fig.height=8, echo=F, fig.align='center'}
pl_scatter_list$species
rm(pl_scatter_list)
invisible(gc())
```


## Taxonomy tables

- Tables with most occurring taxonomies on each rank
- Most occurring is defined in OTU variation and in number of reads
- Note that not all taxons with a lot of reads are also diverse, and vice versa
  - Taxonomic resolution of amplicon can be different ~ taxonomic group

```{r tb_list, echo=F, results='asis'}

for (ttab in tb_list){
  print(ttab)
}

```


# More questions?

You can quite easily load the data into an R session using gendiver:

- https://github.com/inbo/gendiver

Use the following links/paths:

NOTE: you can replace PATHS with the URL of files on the GDRIVE!

`params$SAMPLE_SHEET_URL` = `r params$SAMPLE_SHEET_URL`

`params$OTU_PATH` = `r params$OTU_PATH`

`params$TAX_PATH` = `r params$TAX_PATH`

`params$FASTA_PATH` = `r params$FASTA_PATH`

Examples to load in data (see also INPUT segment of this report):

```
# Load custom R-pkg
library(gendiver)

# read sample sheet
my_sample_data = read.sample_sheet_lab(params$SAMPLE_SHEET_URL)

# read OTU table
my_otu_tab = read.otu_table_vsearch(params$OTU_PATH)

# read taxonomy table (different functions for different tools!)
my_tax_tab = read.taxonomy_sintax(params$TAX_PATH)
my_tax_tab = read.taxonomy_obitools3()

# [optional]  select ID1 taxa from obitools3 data
my_tax_tab = select.taxonomy_obitools3(my_tax_tab)

# read FASTA file
my_seqs = read.fastx_gdrive(params$FASTA_PATH)

```


Or when using the `phyloseq` object created alongside this Rmd (in `r datadir`).

```
my_ps = readRDS(".../data_export/phyloseq.rds")
```
